---
title: "Customer Segmentation: K-means & Hierarchical Clustering"
subtitle: "Applied Predictive & Causal Analytics — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(knitr)

set.seed(42)

policyholders <- read_csv("L2_policyholders.csv")
```

# Overview

This report segments the **`r nrow(policyholders)`-policyholder** portfolio
into distinct customer groups using **k-means** and **hierarchical
clustering** — unsupervised methods that find structure without a target
variable.

Segmentation supports differentiated pricing, targeted retention, and
portfolio risk management.

---

# Step 1 — Prepare and Scale Features

```{r prepare, echo=TRUE}
# Select numeric features for clustering
cluster_features <- policyholders %>%
  select(age, annual_premium, policy_duration_months, claim_amount)

# Scaling is essential — clustering is distance-based
scaled_features <- scale(cluster_features)
```

```{r scale-check}
tibble(
  Feature = colnames(scaled_features),
  `Original Mean` = round(colMeans(cluster_features), 1),
  `Scaled Mean`   = round(colMeans(scaled_features), 3),
  `Scaled SD`     = round(apply(scaled_features, 2, sd), 3)
) %>%
  kable(caption = "Table 1: Effect of Scaling (scaled mean ≈ 0, sd ≈ 1)")
```

---

# Step 2 — Choose k with the Elbow Method

```{r elbow, echo=TRUE}
wss <- map_dbl(1:10, function(k) {
  kmeans(scaled_features, centers = k, nstart = 25)$tot.withinss
})
```

```{r elbow-plot, fig.width=8, fig.height=4, fig.cap="Figure 1: Elbow Method — Within-Cluster Sum of Squares"}
tibble(k = 1:10, wss = wss) %>%
  ggplot(aes(x = k, y = wss)) +
  geom_line(color = "#2c7bb6", linewidth = 0.9) +
  geom_point(size = 2.5, color = "#2c7bb6") +
  geom_vline(xintercept = 4, linetype = "dashed", color = "#d7191c") +
  scale_x_continuous(breaks = 1:10) +
  labs(title    = "Elbow Method for Optimal k",
       subtitle = "Red line = chosen k = 4",
       x = "Number of Clusters (k)",
       y = "Total Within-Cluster Sum of Squares") +
  theme_minimal(base_size = 12)
```

---

# Step 3 — K-means Clustering

```{r kmeans, echo=TRUE}
k <- 4
km_result <- kmeans(scaled_features, centers = k, nstart = 25)

# Attach cluster labels
policyholders$cluster <- factor(km_result$cluster)
```

## Cluster Profiles

```{r cluster-profile}
policyholders %>%
  group_by(Cluster = cluster) %>%
  summarise(
    Size          = n(),
    `Avg Age`     = round(mean(age), 1),
    `Avg Premium` = round(mean(annual_premium), 0),
    `Avg Tenure`  = round(mean(policy_duration_months), 1),
    `Avg Claim`   = round(mean(claim_amount), 0),
    .groups = "drop"
  ) %>%
  kable(caption = "Table 2: K-means Cluster Profiles")
```

## Cluster Visualisation

```{r cluster-plot, fig.width=8, fig.height=5, fig.cap="Figure 2: Clusters by Age and Annual Premium"}
policyholders %>%
  ggplot(aes(x = age, y = annual_premium, color = cluster)) +
  geom_point(alpha = 0.6, size = 1.8) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  labs(title = "Customer Segments by Age and Premium",
       x = "Age", y = "Annual Premium (£)",
       color = "Cluster") +
  theme_minimal(base_size = 12)
```

---

# Step 4 — Hierarchical Clustering

```{r hclust, echo=TRUE}
# Use a sample for a readable dendrogram
set.seed(42)
sample_idx <- sample(nrow(scaled_features), 50)
sample_scaled <- scaled_features[sample_idx, ]

# Distance matrix + linkage
dist_matrix <- dist(sample_scaled, method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
```

```{r dendrogram, fig.width=9, fig.height=5, fig.cap="Figure 3: Hierarchical Clustering Dendrogram (50 sampled customers)"}
plot(hc, labels = FALSE, hang = -1,
     main = "Hierarchical Clustering Dendrogram (Ward linkage)",
     xlab = "Customers", ylab = "Height (dissimilarity)",
     sub = "")
rect.hclust(hc, k = 4, border = c("#d7191c", "#2c7bb6",
                                   "#1a9641", "#fdae61"))
```

## Compare Cluster Assignments

```{r hclust-compare}
hc_clusters <- cutree(hc, k = 4)

tibble(
  `Hierarchical Cluster` = factor(hc_clusters)
) %>%
  count(`Hierarchical Cluster`) %>%
  rename(`Customers (in sample)` = n) %>%
  kable(caption = "Table 3: Hierarchical Cluster Sizes (50-customer sample)")
```

---

# Step 5 — Segment Naming & Interpretation

```{r segment-naming}
segment_summary <- policyholders %>%
  group_by(cluster) %>%
  summarise(
    size           = n(),
    avg_premium    = mean(annual_premium),
    avg_tenure     = mean(policy_duration_months),
    avg_claim      = mean(claim_amount),
    .groups = "drop"
  ) %>%
  mutate(
    segment_label = case_when(
      avg_premium > median(avg_premium) & avg_tenure > median(avg_tenure)
        ~ "High-Value Loyal",
      avg_premium > median(avg_premium) & avg_tenure <= median(avg_tenure)
        ~ "High-Value At-Risk",
      avg_premium <= median(avg_premium) & avg_tenure > median(avg_tenure)
        ~ "Budget Loyal",
      TRUE
        ~ "Budget New/At-Risk"
    )
  )
```

```{r segment-table}
segment_summary %>%
  mutate(
    avg_premium = paste0("£", round(avg_premium, 0)),
    avg_tenure  = round(avg_tenure, 1),
    avg_claim   = paste0("£", round(avg_claim, 0))
  ) %>%
  select(Cluster = cluster, Size = size,
         `Avg Premium` = avg_premium,
         `Avg Tenure` = avg_tenure,
         `Avg Claim` = avg_claim,
         `Segment Label` = segment_label) %>%
  kable(caption = "Table 4: Named Customer Segments")
```

---

# Key Findings

- **Scaling is essential** before clustering — without it, `annual_premium`
  (range in thousands) would dominate `age` (range in tens), distorting
  all distance calculations.
- The **elbow method** (Figure 1) guides the choice of k by identifying
  where adding clusters stops meaningfully reducing within-cluster variance.
- **K-means** assigns every customer to one of k segments; **hierarchical
  clustering** builds a tree (dendrogram) showing how segments nest,
  letting the analyst choose the cut level after seeing the structure.
- The named segments (Table 4) translate abstract clusters into
  actionable customer types — directly usable for differentiated pricing
  and retention strategy.
- In **Lesson 25** — the final lesson — we apply **A/B testing and causal
  inference** to evaluate whether retention interventions actually work.

---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
