---
title: "Motor Insurance Portfolio Review"
subtitle: "Advanced R for Actuarial Statistics — Lesson 2 Business Case"
author: "Actuarial Analytics Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: flatly
    number_sections: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE
)
```

------------------------------------------------------------------------

# Overview

This report presents a **Q4 2024 portfolio review** for the motor insurance book. We examine three policy classes — Comprehensive, Third Party Only, and Fire & Theft — across key actuarial metrics including:

-   Premium volume and exposure
-   Claim frequency and average claim cost
-   Loss ratio by class and region
-   Risk segmentation by vehicle age and policyholder age

> **Data source:** `motor_portfolio_q4_2024.csv` — 100 individual policy records generated for training purposes. Place this file in your working directory before knitting.

------------------------------------------------------------------------

# Data Loading & Preparation

## Load the Dataset

```{r load-data}
# ── Load required packages ──────────────────────────────────────────────────
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

# ── Read the CSV ─────────────────────────────────────────────────────────────
# Make sure motor_portfolio_q4_2024.csv is in your working directory
motor <- read.csv("~/Documents/GitHub/2nd_ifoa/data/motor_portfolio_q4_2024.csv", stringsAsFactors = FALSE)

# Quick check
glimpse(motor)
```

## Data Dictionary

| Column | Type | Description |
|------------------------|------------------------|------------------------|
| `policy_id` | character | Unique policy identifier |
| `policy_class` | character | Comprehensive / Third Party Only / Fire & Theft |
| `region` | character | UK region |
| `vehicle_age` | integer | Age of insured vehicle (years) |
| `annual_premium_gbp` | numeric | Annual gross written premium (£) |
| `claim_made` | integer | 1 = claim made, 0 = no claim |
| `claim_cost_gbp` | numeric | Settled claim amount (£); 0 if no claim |
| `exposure_months` | integer | Policy exposure in months (all = 12 here) |
| `policyholder_age` | integer | Age of primary policyholder |
| `no_claims_years` | integer | Years of continuous no-claims history |

------------------------------------------------------------------------

# Portfolio Summary

## Premium & Exposure by Class

```{r summary-by-class}
# ── Aggregate by policy class ─────────────────────────────────────────────
class_summary <- motor %>%
  group_by(policy_class) %>%
  summarise(
    n_policies      = n(),
    total_premium   = sum(annual_premium_gbp),
    avg_premium     = round(mean(annual_premium_gbp), 0),
    n_claims        = sum(claim_made),
    claim_frequency = round(mean(claim_made), 3),
    avg_claim_cost  = round(mean(claim_cost_gbp[claim_made == 1]), 0),
    .groups = "drop"
  )

kable(class_summary,
      caption   = "Table 1: Portfolio Summary by Policy Class",
      col.names = c("Class", "Policies", "Total Premium (£)",
                    "Avg Premium (£)", "Claims", "Claim Freq", "Avg Claim Cost (£)"),
      format.args = list(big.mark = ","))
```

## Regional Distribution

```{r region-table}
# ── Count policies per region ─────────────────────────────────────────────
region_summary <- motor %>%
  group_by(region) %>%
  summarise(
    n_policies    = n(),
    avg_premium   = round(mean(annual_premium_gbp), 0),
    claim_freq    = round(mean(claim_made), 3),
    .groups = "drop"
  ) %>%
  arrange(desc(n_policies))

kable(region_summary,
      caption   = "Table 2: Portfolio Distribution by Region",
      col.names = c("Region", "Policies", "Avg Premium (£)", "Claim Freq"))
```

------------------------------------------------------------------------

# Loss Ratio Analysis

## Calculating the Loss Ratio

The **loss ratio** is a fundamental actuarial metric defined as:

$$\text{Loss Ratio} = \frac{\text{Total Claims Incurred}}{\text{Total Premiums Earned}}$$

A loss ratio above **1.0 (100%)** indicates that claims exceed premiums — a signal of underpricing or adverse claims experience.

```{r loss-ratio}
# ── Loss ratio by class ───────────────────────────────────────────────────
loss_ratio <- motor %>%
  group_by(policy_class) %>%
  summarise(
    total_premium = sum(annual_premium_gbp),
    total_claims  = sum(claim_cost_gbp),
    loss_ratio    = round(total_claims / total_premium, 3),
    .groups = "drop"
  )

# ── Identify highest-risk class ───────────────────────────────────────────
worst_class <- loss_ratio$policy_class[which.max(loss_ratio$loss_ratio)]
worst_lr    <- max(loss_ratio$loss_ratio)

kable(loss_ratio,
      caption   = "Table 3: Loss Ratio by Policy Class",
      col.names = c("Class", "Total Premium (£)", "Total Claims (£)", "Loss Ratio"),
      format.args = list(big.mark = ","))
```

The policy class with the **highest loss ratio** is **`r worst_class`**, at **`r worst_lr`**. This class warrants attention in the next pricing review cycle.

------------------------------------------------------------------------

# Visualisations

## Loss Ratio Chart

```{r fig-loss-ratio, fig.cap="Figure 1: Loss Ratio by Policy Class", fig.width=7, fig.height=4}
ggplot(loss_ratio, aes(x = reorder(policy_class, -loss_ratio),
                        y = loss_ratio, fill = policy_class)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 1.0, linetype = "dashed",
             colour = "red", linewidth = 0.8) +
  geom_text(aes(label = paste0(loss_ratio * 100, "%")),
            vjust = -0.4, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Comprehensive"    = "#1F4E79",
                                "Third Party Only" = "#2E86AB",
                                "Fire & Theft"     = "#A8DADC")) +
  labs(
    title    = "Loss Ratio by Policy Class — Q4 2024",
    subtitle = "Red dashed line = break-even (100%)",
    x        = "Policy Class",
    y        = "Loss Ratio"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title       = element_text(face = "bold"))
```

## Claim Frequency by Region

```{r fig-region, fig.cap="Figure 2: Claim Frequency by Region", fig.width=7, fig.height=4}
ggplot(region_summary, aes(x = reorder(region, claim_freq),
                            y = claim_freq, fill = claim_freq)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "#BDD7EE", high = "#1F4E79") +
  labs(
    title = "Claim Frequency by Region",
    x     = NULL,
    y     = "Claim Frequency"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## Premium Distribution

```{r fig-premium-dist, fig.cap="Figure 3: Premium Distribution by Policy Class", fig.width=7, fig.height=4}
ggplot(motor, aes(x = annual_premium_gbp, fill = policy_class)) +
  geom_histogram(bins = 20, alpha = 0.8, colour = "white") +
  facet_wrap(~ policy_class, scales = "free_x") +
  scale_fill_manual(values = c("Comprehensive"    = "#1F4E79",
                                "Third Party Only" = "#2E86AB",
                                "Fire & Theft"     = "#A8DADC")) +
  labs(
    title = "Premium Distribution by Policy Class",
    x     = "Annual Premium (£)",
    y     = "Number of Policies"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

------------------------------------------------------------------------

# Risk Segmentation

## ★ STUDENT EXERCISE — Complete this section during the lesson

The section below is **intentionally incomplete**. During the lesson, you will:

1.  Calculate the average loss ratio by **vehicle age band** (0–2 yrs, 3–5 yrs, 6+ yrs)
2.  Produce a scatter plot of `policyholder_age` vs `annual_premium_gbp`, coloured by `claim_made`
3.  Write one sentence of **inline R interpretation** using the result

```{r risk-segmentation, eval=FALSE}
# ── STEP 1: Create vehicle age bands ─────────────────────────────────────
motor <- motor %>%
  mutate(
    age_band = case_when(
      vehicle_age <= 2 ~ "0-2 years",
      vehicle_age <= 5 ~ "3-5 years",
      TRUE             ~ "6+ years"
    )
  )

# ── STEP 2: Your code here ────────────────────────────────────────────────
# Group by age_band and calculate: n_policies, avg_premium, claim_freq, loss_ratio
# Hint: use group_by() and summarise()

age_band_summary <- motor %>%
  group_by(___) %>%
  summarise(
    n_policies    = n(),
    avg_premium   = ___,
    claim_freq    = ___,
    total_premium = ___,
    total_claims  = ___,
    loss_ratio    = ___
  )

kable(age_band_summary, caption = "Table 4: Loss Ratio by Vehicle Age Band")


# ── STEP 3: Scatter plot ──────────────────────────────────────────────────
ggplot(motor, aes(x = ___, y = ___, colour = factor(claim_made))) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "Policyholder Age vs Annual Premium",
       colour = "Claim Made")
```

------------------------------------------------------------------------

# Key Findings

```{r key-findings}
# ── Summary statistics for narrative ─────────────────────────────────────
total_policies  <- nrow(motor)
total_premium   <- sum(motor$annual_premium_gbp)
total_claims_n  <- sum(motor$claim_made)
overall_freq    <- round(mean(motor$claim_made), 3)
overall_lr      <- round(sum(motor$claim_cost_gbp) / sum(motor$annual_premium_gbp), 3)
```

Based on the Q4 2024 portfolio analysis:

-   The portfolio comprises **`r total_policies` policies** with total GWP of **£`r format(total_premium, big.mark=",")`**.
-   **`r total_claims_n` claims** were reported, giving an overall claim frequency of **`r overall_freq * 100`%**.
-   The portfolio-level loss ratio stands at **`r overall_lr * 100`%**.
-   `r worst_class` has the highest loss ratio at `r worst_lr * 100`%, suggesting a pricing review may be warranted.

------------------------------------------------------------------------

*Report generated automatically using R Markdown. Re-knit with updated data to refresh all figures and narrative.*
