---
title: "EDA Report: Policyholder Churn & Retention"
subtitle: "Exploratory Data Analysis — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(knitr)
library(skimr)

policyholders <- read_csv("policyholders.csv") %>%
  mutate(churned = if_else(policy_duration_months < 24, 1, 0))
```

# Dataset Overview

This report explores churn patterns across the policyholder portfolio.
The dataset contains **`r nrow(policyholders)`** policyholders.
Overall, **`r sum(policyholders$churned)`** policyholders churned
(**`r paste0(round(mean(policyholders$churned)*100, 1), "%")`** churn rate).

---

# Variable Summary

```{r skim, echo=TRUE}
policyholders %>%
  select(age, annual_premium, policy_duration_months, churned) %>%
  skim()
```

---

# Churn Rate by Policy Type

```{r churn-by-type, echo=TRUE}
churn_by_type <- policyholders %>%
  group_by(policy_type) %>%
  summarise(
    Policyholders = n(),
    Churned       = sum(churned),
    `Churn Rate`  = paste0(round(mean(churned) * 100, 1), "%")
  )
```

```{r churn-type-table}
churn_by_type %>%
  kable(caption = "Table 1: Churn Rate by Policy Type")
```

```{r churn-type-plot, echo=FALSE, fig.width=7, fig.height=4,
    fig.cap="Figure 1: Churn Rate by Policy Type"}
policyholders %>%
  group_by(policy_type) %>%
  summarise(churn_rate = mean(churned) * 100) %>%
  ggplot(aes(x = reorder(policy_type, churn_rate),
             y = churn_rate,
             fill = policy_type)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = paste0(round(churn_rate, 1), "%")),
            hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(limits = c(0, 80)) +
  labs(title = "Churn Rate by Policy Type",
       x = NULL, y = "Churn Rate (%)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

# Churn Rate by Region

```{r churn-by-region, echo=FALSE, fig.width=7, fig.height=4,
    fig.cap="Figure 2: Churn Rate by Region"}
policyholders %>%
  group_by(region) %>%
  summarise(churn_rate = mean(churned) * 100) %>%
  ggplot(aes(x = reorder(region, churn_rate),
             y = churn_rate,
             fill = churn_rate)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = paste0(round(churn_rate, 1), "%")),
            hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low = "#ffffb2", high = "#d7191c") +
  scale_y_continuous(limits = c(0, 80)) +
  labs(title = "Churn Rate by Region",
       x = NULL, y = "Churn Rate (%)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

# Age Distribution by Churn Status

```{r age-churn, echo=FALSE, fig.width=7, fig.height=4,
    fig.cap="Figure 3: Age Distribution by Churn Status"}
policyholders %>%
  mutate(Status = if_else(churned == 1, "Churned", "Retained")) %>%
  ggplot(aes(x = age, fill = Status)) +
  geom_histogram(binwidth = 5, alpha = 0.7,
                 position = "identity", color = "white") +
  scale_fill_manual(values = c("Churned" = "#d7191c",
                                "Retained" = "#2c7bb6")) +
  labs(title = "Age Distribution by Churn Status",
       x = "Age", y = "Count") +
  theme_minimal(base_size = 12) +
  facet_wrap(~Status)
```

---

# Key Findings

- Overall churn rate: **`r paste0(round(mean(policyholders$churned)*100,1),"%")`**
- Churn varies meaningfully across product lines and regions — warranting deeper modelling.
- Age distributions differ between churned and retained policyholders, suggesting age may be a predictive feature.

---

*Report generated automatically using R Markdown. Data source: `policyholders.csv`.*
