---
title: "Time Series Objects: ts and xts"
subtitle: "Time Series Basics — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(xts)
library(forecast)
library(knitr)

monthly_premiums <- read_csv("L5_monthly_premiums.csv") %>%
  mutate(month = as.Date(month))

# Build ts object
premium_ts <- ts(
  monthly_premiums$total_premium,
  start     = c(2019, 1),
  frequency = 12
)

# Build xts object
premium_xts <- xts(
  monthly_premiums$total_premium,
  order.by = monthly_premiums$month
)
```

# Overview

This report demonstrates the structure and practical use of `ts` and `xts`
objects using **`r nrow(monthly_premiums)` months** of premium income data,
running from **`r format(min(monthly_premiums$month), '%B %Y')`** to
**`r format(max(monthly_premiums$month), '%B %Y')`**.

---

# The ts Object

## Structure

```{r ts-str, echo=TRUE}
# Inspect the ts object
print(premium_ts)
```

```{r ts-attributes, echo=TRUE}
# Key attributes
cat("Start:    ", start(premium_ts), "\n")
cat("End:      ", end(premium_ts),   "\n")
cat("Frequency:", frequency(premium_ts), "\n")
cat("Length:   ", length(premium_ts), "\n")
```

## Indexing a ts Object

```{r ts-index, echo=TRUE}
# Extract a specific window
window(premium_ts, start = c(2021, 1), end = c(2021, 12))
```

## Visualisation

```{r ts-plot, echo=FALSE, fig.width=8, fig.height=4,fig.cap="Figure 1: Monthly Premium Income — ts Object"}
autoplot(premium_ts) +
  labs(title = "Monthly Premium Income (ts)",
       x = NULL, y = "Total Premium (£)") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  theme_minimal(base_size = 12)
```

---

# The xts Object

## Structure

```{r xts-str, echo=TRUE}
# Inspect the xts object
head(premium_xts, 6)
```

```{r xts-periodicity, echo=TRUE}
# Check periodicity
periodicity(premium_xts)
```

## Indexing an xts Object

```{r xts-index, echo=TRUE}
# Index by year
premium_xts["2021"]
```

```{r xts-range, echo=TRUE}
# Index by date range
premium_xts["2021-01/2021-06"]
```

## Year-End Values with endpoints()

```{r xts-endpoints, echo=TRUE}
# Extract the last value of each year
ep <- endpoints(premium_xts, on = "years")
premium_xts[ep]
```

## Visualisation

```{r xts-plot, echo=FALSE, fig.width=8, fig.height=4,fig.cap="Figure 2: Monthly Premium Income — xts Object"}
autoplot(premium_xts) +
  labs(title = "Monthly Premium Income (xts)",
       x = NULL, y = "Total Premium (£)") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  theme_minimal(base_size = 12)
```

---

# Converting Between ts and xts

```{r convert, echo=TRUE}
# ts → xts
converted_to_xts <- as.xts(premium_ts)

# xts → ts
converted_to_ts  <- as.ts(premium_xts)

cat("Converted xts class:", class(converted_to_xts), "\n")
cat("Converted ts class: ", class(converted_to_ts),  "\n")
```

---

# Annual Summary Table

```{r annual-summary}
monthly_premiums %>%
  mutate(year = lubridate::year(month)) %>%
  group_by(Year = year) %>%
  summarise(
    `Months`          = n(),
    `Min Premium (£)` = round(min(total_premium), 0),
    `Avg Premium (£)` = round(mean(total_premium), 0),
    `Max Premium (£)` = round(max(total_premium), 0),
    `Total GWP (£)`   = format(round(sum(total_premium), 0), big.mark = ",") #1,234,567
  ) %>%
  kable(caption = "Table 1: Annual Premium Summary")
```

---

# Key Findings

- The `ts` object is best suited for **regular, equally-spaced** series where
  frequency is fixed (monthly, quarterly, annual). It integrates directly with
  `forecast`, `decompose`, and `arima` functions used in Lessons 7–9.
- The `xts` object uses **actual calendar dates** as its index, making it more
  flexible for irregular series and date-based subsetting.
- Both formats can be converted with `as.xts()` and `as.ts()`, so the choice
  of format depends on the downstream analysis required.

---

*Report generated automatically using R Markdown.
Data source: `monthly_premiums.csv`.*
