---
title: "ARIMA Modelling of Monthly Premium Income"
subtitle: "Time Series Modelling — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(forecast)
library(knitr)

monthly_premiums <- read_csv("L5_monthly_premiums.csv") %>%
  mutate(month = as.Date(month))

premium_ts <- ts(
  monthly_premiums$total_premium,
  start     = c(2019, 1),
  frequency = 12
)
```

# Overview

This report fits an ARIMA model to the monthly premium income series
analysed in Lesson 7. The series runs from
**`r format(min(monthly_premiums$month), '%B %Y')`** to
**`r format(max(monthly_premiums$month), '%B %Y')`**
(**`r length(premium_ts)`** observations).

Lesson 7 diagnostics indicated: slow ACF decay suggesting differencing
is required, and PACF cutting off after lag 1 suggesting an AR(1)
structure. These findings guide the model specification below.

---

# Step 1 — Determine Differencing Order

```{r ndiffs, echo=TRUE}
d  <- ndiffs(premium_ts)
ds <- nsdiffs(premium_ts)

cat("Non-seasonal differences required:", d,  "\n")
cat("Seasonal differences required:    ", ds, "\n")
```

---

# Step 2 — Automatic Model Selection

```{r auto-arima, echo=TRUE}
model_auto <- auto.arima(
  premium_ts,
  stepwise   = FALSE,
  approximation = FALSE
)

summary(model_auto)
```

```{r auto-arima-order}
ord <- arimaorder(model_auto)
cat("Selected order: ARIMA(",
    ord["p"], ",", ord["d"], ",", ord["q"], ")",
    "× (", ord["P"], ",", ord["D"], ",", ord["Q"], ")[12]\n")
```

---

# Step 3 — Manual Model for Comparison

```{r manual-arima, echo=TRUE}
model_manual <- Arima(
  premium_ts,
  order    = c(1, 1, 0),
  seasonal = c(1, 1, 0)
)

summary(model_manual)
```

---

# Step 4 — Model Comparison

```{r model-comparison}
tibble(
  Model = c(
    paste0("auto.arima — ARIMA(",
           arimaorder(model_auto)["p"], ",",
           arimaorder(model_auto)["d"], ",",
           arimaorder(model_auto)["q"], ")(",
           arimaorder(model_auto)["P"], ",",
           arimaorder(model_auto)["D"], ",",
           arimaorder(model_auto)["Q"], ")[12]"),
    "Manual — ARIMA(1,1,0)(1,1,0)[12]"
  ),
  AIC  = round(c(model_auto$aic,   model_manual$aic),   2),
  BIC  = round(c(model_auto$bic,   model_manual$bic),   2),
  RMSE = round(c(accuracy(model_auto)[, "RMSE"],
                  accuracy(model_manual)[, "RMSE"]), 2)
) %>%
  kable(caption = "Table 1: Model Comparison — AIC, BIC, RMSE")
```

---

# Step 5 — Residual Diagnostics

## Auto ARIMA Residuals

```{r residuals-auto, echo=TRUE, fig.width=8, fig.height=5,fig.cap="Figure 1: Residual Diagnostics — auto.arima Model"}
checkresiduals(model_auto)
```

## Ljung-Box Test

```{r ljung-box, echo=TRUE}
lb_test <- Box.test(residuals(model_auto),
                    lag    = 24,
                    type   = "Ljung-Box",
                    fitdf  = sum(arimaorder(model_auto)[c("p","q","P","Q")]))

cat("Ljung-Box p-value (lag 24):", round(lb_test$p.value, 4), "\n")
cat("Interpretation:",
    ifelse(lb_test$p.value > 0.05,
           "Residuals are consistent with white noise.",
           "Residuals show remaining autocorrelation — model may need revision."),
    "\n")
```

---

# Key Findings

- `ndiffs()` confirmed **`r ndiffs(premium_ts)` non-seasonal difference** and
  `nsdiffs()` confirmed **`r nsdiffs(premium_ts)` seasonal difference** required,
  consistent with the slow ACF decay observed in Lesson 7.
- `auto.arima()` selected the model with the lowest AIC across a full grid
  search (`stepwise = FALSE`).
- The Ljung-Box test p-value of
  **`r round(Box.test(residuals(model_auto), lag=24, type='Ljung-Box', fitdf=sum(arimaorder(model_auto)[c('p','q','P','Q')]))$p.value, 4)`**
  `r ifelse(Box.test(residuals(model_auto), lag=24, type='Ljung-Box', fitdf=sum(arimaorder(model_auto)[c('p','q','P','Q')]))$p.value > 0.05, "confirms residuals are consistent with white noise — the model is adequate.", "suggests residuals retain structure — further refinement may be needed.")`
- The fitted model is carried forward into **Lesson 9** for forecasting
  and forecast interval construction.

---

*Report generated automatically using R Markdown.
Data source: `monthly_premiums.csv`.*
