---
title: "Markov Chains: Policyholder State Transitions"
subtitle: "Stochastic Modelling — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(markovchain)
library(knitr)
library(purrr)
library(tibble)
library(expm)


set.seed(42)

policyholders <- read_csv("L2_policyholders.csv")
```

# Overview

This report models policyholder status — **Active**, **Lapsed**, and
**Claimed** — as a discrete-time Markov chain, calibrated from the
**`r nrow(policyholders)`-policyholder** portfolio.

A Markov chain is appropriate here because we assume next period's
status depends only on the current period's status — not on the full
history of previous states. This is the standard actuarial assumption
for lapse and claim modelling.

---

# Transition Matrix

## Empirical Transitions from Portfolio Data

```{r empirical-transitions, echo=TRUE}
# Derive state from portfolio data
portfolio <- policyholders %>%
  mutate(state = case_when(
    has_claim == 1                      ~ "Claimed",
    policy_duration_months < 24         ~ "Lapsed",
    TRUE                                ~ "Active"
  ))

# Empirical transition counts
trans_counts <- portfolio %>%
  count(state) %>%
  mutate(total = sum(n), prop = n / total)
```

```{r counts-table}
trans_counts %>%
  mutate(prop = paste0(round(prop * 100, 1), "%")) %>%
  rename(State = state, Count = n, `Portfolio Share` = prop) %>%
  select(State, Count, `Portfolio Share`) %>%
  kable(caption = "Table 1: Observed Portfolio State Distribution")
```

## Transition Matrix Definition

```{r transition-matrix, echo=TRUE}
# Define 3-state transition probability matrix
# Rows: current state | Columns: next state
P <- matrix(
  c(0.85, 0.10, 0.05,   # Active   → Active, Lapsed, Claimed
    0.00, 1.00, 0.00,   # Lapsed   → Active, Lapsed, Claimed
    0.00, 0.00, 1.00),  # Claimed  → Active, Lapsed, Claimed
  nrow  = 3,
  byrow = TRUE,
  dimnames = list(
    c("Active", "Lapsed", "Claimed"),
    c("Active", "Lapsed", "Claimed")
  )
)
```

```{r matrix-table}
as.data.frame(P) %>%
  rownames_to_column("From \\ To") %>%
  kable(caption = "Table 2: Transition Probability Matrix",
        digits = 2)
```

---

# Markov Chain Object

```{r mc-object, echo=TRUE}
mc <- new("markovchain",
          states       = c("Active", "Lapsed", "Claimed"),
          transitionMatrix = P,
          name         = "Policyholder Status")

mc
```

---

# Multi-Step Transitions

## State Distribution After n Steps

```{r nstep, echo=TRUE}
# Initial state: 100% Active
initial <- c(Active = 1, Lapsed = 0, Claimed = 0)

# State distribution after 1, 6, 12, 24 months
steps <- c(1, 6, 12, 24)

step_results <- map_dfr(steps, function(n) {
  dist <- initial %*% (P %^% n)
  tibble(
    Months  = n,
    Active  = round(dist[1, "Active"],  4),
    Lapsed  = round(dist[1, "Lapsed"],  4),
    Claimed = round(dist[1, "Claimed"], 4)
  )
})
```

```{r nstep-table}
step_results %>%
  mutate(across(c(Active, Lapsed, Claimed),
                ~ paste0(round(. * 100, 1), "%"))) %>%
  kable(caption = "Table 3: State Distribution Over Time (Starting 100% Active)")
```

```{r nstep-plot, fig.width=8, fig.height=4, fig.cap="Figure 1: State Distribution Over 36 Months"}
map_dfr(1:36, function(n) {
  dist <- initial %*% (P %^% n)
  tibble(Month   = n,
         Active  = dist[1, "Active"],
         Lapsed  = dist[1, "Lapsed"],
         Claimed = dist[1, "Claimed"])
}) %>%
  pivot_longer(-Month, names_to = "State", values_to = "Proportion") %>%
  ggplot(aes(x = Month, y = Proportion, color = State)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("Active"  = "#2c7bb6",
                                 "Lapsed"  = "#d7191c",
                                 "Claimed" = "#fdae61")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "State Distribution Over 36 Months",
       x = "Month", y = "Proportion of Policyholders") +
  theme_minimal(base_size = 12)
```

---

# Stationary Distribution

```{r steady-state, echo=TRUE}
ss <- steadyStates(mc)
```

```{r steady-state-table}
as.data.frame(ss) %>%
  mutate(across(everything(), ~ paste0(round(. * 100, 1), "%"))) %>%
  kable(caption = "Table 4: Long-Run Stationary Distribution")
```

The stationary distribution represents the long-run proportion of
policyholders in each state, given the current transition probabilities.
Since **Lapsed** and **Claimed** are absorbing states, all policyholders
eventually leave the Active state — the stationary distribution reflects this.

---

# Simulation: 1,000 Policyholder Paths

```{r simulation, echo=TRUE}
n_policyholders <- 1000
n_months        <- 36

sim_paths <- replicate(n_policyholders, {
  path    <- character(n_months)
  current <- "Active"
  for (t in 1:n_months) {
    probs   <- P[current, ]
    current <- sample(colnames(P), size = 1, prob = probs)
    path[t] <- current
  }
  path
})
```

```{r sim-summary}
# Month-by-month state counts
month_dist <- map_dfr(1:n_months, function(t) {
  states <- sim_paths[t, ]
  tibble(
    Month   = t,
    Active  = mean(states == "Active"),
    Lapsed  = mean(states == "Lapsed"),
    Claimed = mean(states == "Claimed")
  )
})
```

```{r sim-plot, fig.width=8, fig.height=4,fig.cap="Figure 2: Simulated State Distribution — 1,000 Policyholder Paths"}
month_dist %>%
  pivot_longer(-Month, names_to = "State", values_to = "Proportion") %>%
  ggplot(aes(x = Month, y = Proportion, color = State)) +
  geom_line(linewidth = 0.8, alpha = 0.9) +
  scale_color_manual(values = c("Active"  = "#2c7bb6",
                                 "Lapsed"  = "#d7191c",
                                 "Claimed" = "#fdae61")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Simulated State Distribution Over 36 Months",
       subtitle = "1,000 independent policyholder paths",
       x = "Month", y = "Proportion of Policyholders") +
  theme_minimal(base_size = 12)
```

---

# Key Findings

```{r kf-vals, include=FALSE}
active_12  <- step_results %>% filter(Months == 12) %>% pull(Active)
lapsed_12  <- step_results %>% filter(Months == 12) %>% pull(Lapsed)
claimed_12 <- step_results %>% filter(Months == 12) %>% pull(Claimed)
```

- Starting from a fully active cohort, after **12 months**:
  **`r paste0(round(active_12*100,1),"%")`** remain Active,
  **`r paste0(round(lapsed_12*100,1),"%")`** have Lapsed, and
  **`r paste0(round(claimed_12*100,1),"%")`** have Claimed.
- **Lapsed** and **Claimed** are absorbing states — once a policyholder
  enters either state, they cannot return to Active under this model.
- The analytical multi-step result (Table 3) and the simulation result
  (Figure 2) converge — confirming the simulation is correctly implemented.
- In **Lesson 12**, we introduce the Poisson process to model the
  *number* of claims arriving over time, complementing the state-based
  view developed here.

---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
