---
title: "Policyholder Portfolio Summary"
subtitle: "Monthly Reporting — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(knitr)

policyholders <- read_csv("policyholders.csv")
```

# Executive Summary

The portfolio contains **`r nrow(policyholders)`** policyholders
across **`r n_distinct(policyholders$policy_type)`** product lines.
The average annual premium is **£`r round(mean(policyholders$annual_premium), 0)`**.
The overall claim rate is **`r paste0(round(mean(policyholders$has_claim) * 100, 1), "%")`**.

---

# Claim Experience

## Claim Frequency by Policy Type

```{r claim-summary, echo=FALSE}
policyholders %>%
  group_by(policy_type) %>%
  summarise(
    Policyholders = n(),
    Claims        = sum(has_claim),
    `Claim Rate`  = paste0(round(mean(has_claim) * 100, 1), "%")
  ) %>%
  kable(caption = "Table 1: Claim Frequency by Policy Type")
```

---

*Report generated automatically using R Markdown. Data source: `policyholders.csv`.*
