---
title: "Policy Data Cleaning & Preparation"
subtitle: "Advanced R for Actuarial Statistics — Lesson 3 Business Case"
author: "Actuarial Analytics Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: flatly
    number_sections: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Overview

Real actuarial data is rarely clean. This business case works through a realistic
**policy administration extract** — 25 records from a motor insurance portfolio —
that contains every data quality problem you will commonly encounter in practice:

- Inconsistent date formats (`15/03/1985`, `2024-01-01`, `Jan 5 2024`, ...)
- Mixed-case and whitespace issues in text fields
- Non-standard premium values (`£852.00` as a string)
- Free-text notes fields requiring extraction

We will clean this dataset using three packages:

| Package | Purpose |
|---|---|
| `tidyr` | Reshaping and structural cleaning |
| `stringr` | String detection, replacement, extraction |
| `lubridate` | Date parsing and arithmetic |

> **Data file:** `lesson3_raw_policies.csv` — place in your working directory before knitting.

---

# Load & Inspect

```{r load}
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)

raw <- read.csv("lesson3_raw_policies.csv",
                stringsAsFactors = FALSE,
                na.strings        = c("", "NA", "N/A"))

# Dimensions
dim(raw)

# Column types as-read
glimpse(raw)
```

## Initial Quality Flags

```{r quality-flags}
# Quick count of NAs per column
na_counts <- colSums(is.na(raw))
na_counts[na_counts > 0]
```

At this stage every date column is `chr` — R has no idea they are dates yet.
The `premium_gbp` column contains pound signs, making it character not numeric.
`cover_type` has at least five different spellings of the same three values.

---

# Step 1 — Clean Policyholder Names (stringr)

Names arrive in inconsistent case and with extra whitespace.

```{r clean-names}
policies <- raw %>%
  mutate(
    # Collapse multiple internal spaces to one, strip leading/trailing
    policyholder_name = str_squish(policyholder_name),

    # Convert to title case: "john smith" → "John Smith"
    policyholder_name = str_to_title(policyholder_name)
  )

# Verify
policies %>% select(policy_ref, policyholder_name) %>% head(8)
```

**What changed?**
`str_squish()` handles the `"David    Brown"` and `"Peter  Clarke"` cases.
`str_to_title()` fixes both all-caps (`"ROBERT JOHNSON"`) and all-lower (`"jane doe"`) entries.

---

# Step 2 — Standardise Cover Type (stringr)

```{r clean-cover}
policies <- policies %>%
  mutate(
    # Uppercase everything first, then recode to canonical labels
    cover_type = str_to_upper(str_trim(cover_type)),
    cover_type = case_when(
      str_detect(cover_type, "COMP")        ~ "Comprehensive",
      str_detect(cover_type, "THIRD|3RD")   ~ "Third Party Only",
      str_detect(cover_type, "FIRE")        ~ "Fire & Theft",
      TRUE                                  ~ NA_character_
    )
  )

table(policies$cover_type)
```

`str_detect()` uses partial matching, so `"COMPREHENSIVE"`, `"comprehensive"`,
and `"Comprehensive"` all collapse to the same canonical value.

---

# Step 3 — Clean Region Codes (stringr)

```{r clean-region}
policies <- policies %>%
  mutate(
    region_code = str_trim(region_code)   # removes leading/trailing spaces
  )

sort(unique(policies$region_code))
```

Several region codes had invisible trailing spaces (`"LDN "`, `" SE"`) that would
cause silent grouping errors downstream. `str_trim()` is a one-liner fix.

---

# Step 4 — Parse Premium to Numeric (stringr)

```{r clean-premium}
policies <- policies %>%
  mutate(
    # Remove £ sign and commas, then coerce to numeric
    premium_gbp = str_remove_all(premium_gbp, "[£,]"),
    premium_gbp = as.numeric(premium_gbp)
  )

summary(policies$premium_gbp)
```

`str_remove_all()` with a character class `[£,]` removes every occurrence of
either symbol. After that, `as.numeric()` succeeds cleanly.

---

# Step 5 — Parse Dates (lubridate)

This is usually the hardest step. Our dataset contains **six different date formats**:

```
15/03/1985    (dmy)
22-08-1990    (dmy with dash)
1978.11.05    (ymd with dot)
Jan 5 2024    (mdy text)
2024/01/15    (ymd with slash)
July 1 2024   (mdy text, full month name)
```

lubridate's `parse_date_time()` accepts a vector of format orders and tries each one:

```{r parse-dates}
date_formats <- c("dmy", "ymd", "mdy", "dmy", "ymd", "mdy")

policies <- policies %>%
  mutate(
    dob          = parse_date_time(dob,          orders = date_formats),
    policy_start = parse_date_time(policy_start, orders = date_formats),
    policy_end   = parse_date_time(policy_end,   orders = date_formats),
    claim_date   = parse_date_time(claim_date,   orders = date_formats),

    # Coerce to Date (drops the time component)
    dob          = as.Date(dob),
    policy_start = as.Date(policy_start),
    policy_end   = as.Date(policy_end),
    claim_date   = as.Date(claim_date)
  )

# Verify
policies %>%
  select(policy_ref, dob, policy_start, policy_end, claim_date) %>%
  head(6)
```

---

# Step 6 — Derive Actuarial Variables (lubridate)

Once dates are parsed, we can compute derived fields that are essential for
actuarial analysis.

```{r derive-dates}
reference_date <- as.Date("2025-01-01")

policies <- policies %>%
  mutate(
    # Age at policy start (in whole years)
    age_at_start = as.integer(
      interval(dob, policy_start) / years(1)
    ),

    # Policy duration in days
    policy_duration_days = as.integer(policy_end - policy_start),

    # Days from policy start to claim (NA if no claim)
    days_to_claim = as.integer(claim_date - policy_start),

    # Policy status: Active or Expired as at reference date
    policy_status = if_else(policy_end >= reference_date, "Active", "Expired")
  )

policies %>%
  select(policy_ref, age_at_start, policy_duration_days,
         days_to_claim, policy_status) %>%
  head(8)
```

**Key lubridate functions used:**

| Function | Purpose |
|---|---|
| `interval(start, end)` | Creates a time interval object |
| `years(1)` | Divides interval into whole years |
| `as.integer(date2 - date1)` | Difference in days |
| `if_else()` | Vectorised conditional |

---

# Step 7 — Reshape with tidyr (Bonus)

Sometimes raw data arrives in wide format. Here we demonstrate `pivot_longer()`
by reshaping the date columns into a long format — useful for timeline visualisations.

```{r pivot-longer}
# Select a subset for demonstration
date_long <- policies %>%
  select(policy_ref, policyholder_name,
         policy_start, policy_end, claim_date) %>%
  pivot_longer(
    cols      = c(policy_start, policy_end, claim_date),
    names_to  = "date_type",
    values_to = "date_value"
  ) %>%
  filter(!is.na(date_value)) %>%
  arrange(policy_ref, date_value)

head(date_long, 10)
```

`pivot_longer()` takes multiple columns and stacks them into two:
`date_type` (which column the value came from) and `date_value` (the actual date).
This format is often needed for `ggplot2` timelines and survival analysis inputs.

---

# Step 8 — Extract from Free Text (stringr)

The `notes` field contains unstructured text. We can extract structured signals:

```{r extract-notes}
policies <- policies %>%
  mutate(
    # Flag policies mentioning a number of claims
    multi_claim_flag = str_detect(
      str_to_lower(notes),
      "two claim|multiple|third party only damage"
    ),

    # Extract years of no-claims mentioned in notes (e.g. "4 years")
    ncb_years_noted = str_extract(notes, "\\d+(?= year)")
  )

policies %>%
  filter(!is.na(notes)) %>%
  select(policy_ref, notes, multi_claim_flag, ncb_years_noted) %>%
  head(8)
```

---

# ★ STUDENT EXERCISE

The section below is **intentionally incomplete** (`eval=FALSE`).
Change to `eval=TRUE` and fill in the blanks.

**Tasks:**

1. Create a new column `age_band` grouping `age_at_start` into:
   `"Under 25"`, `"25–39"`, `"40–59"`, `"60+"` using `case_when()`
2. Summarise the cleaned dataset by `cover_type` to show:
   mean premium, mean age at start, claim frequency, mean days to claim
3. Use `str_pad()` to create a formatted policy reference like `"Policy: POL-0001"`

```{r student-exercise, eval=FALSE}
# Task 1 — Age bands
policies <- policies %>%
  mutate(
    age_band = case_when(
      age_at_start < ___          ~ "Under 25",
      age_at_start < ___          ~ "25-39",
      age_at_start < ___          ~ "40-59",
      TRUE                        ~ "60+"
    )
  )

# Task 2 — Summary table by cover type
summary_table <- policies %>%
  group_by(___) %>%
  summarise(
    n              = n(),
    mean_premium   = round(mean(___), 0),
    mean_age       = round(mean(___), 1),
    claim_freq     = round(mean(claim_flag == ___), 3),
    mean_days_claim = round(mean(___, na.rm = TRUE), 0)
  )

knitr::kable(summary_table, caption = "Table: Cleaned Portfolio Summary by Cover Type")

# Task 3 — Formatted reference
policies <- policies %>%
  mutate(
    policy_label = str_c("Policy: ", ___)
  )

head(policies$policy_label)
```

---

# Final Clean Dataset

```{r final-dataset}
# Select the columns we want in the clean output
clean_policies <- policies %>%
  select(
    policy_ref,
    policyholder_name,
    dob,
    age_at_start,
    policy_start,
    policy_end,
    policy_duration_days,
    policy_status,
    cover_type,
    region_code,
    premium_gbp,
    claim_flag,
    claim_date,
    days_to_claim
  ) %>%
  arrange(policy_ref)

# Summary of what we've fixed
cat("Records:", nrow(clean_policies), "\n")
cat("Remaining NAs in key columns:\n")
print(colSums(is.na(clean_policies[, c("cover_type","premium_gbp",
                                        "policy_start","age_at_start")])))
```

## Data Quality Summary

| Issue | Columns Affected | Fix Applied |
|---|---|---|
| Mixed case / extra spaces | `policyholder_name`, `cover_type`, `region_code` | `str_to_title()`, `str_squish()`, `str_trim()` |
| Non-standard cover type labels | `cover_type` | `str_detect()` + `case_when()` |
| Currency symbol in numeric field | `premium_gbp` | `str_remove_all()` + `as.numeric()` |
| Six different date formats | `dob`, `policy_start`, `policy_end`, `claim_date` | `parse_date_time()` |
| Derived variables not present | — | `interval()`, `years()`, date arithmetic |
| Wide date format | Multiple date cols | `pivot_longer()` |
| Unstructured notes | `notes` | `str_detect()`, `str_extract()` |

---

*Report generated automatically using R Markdown. Re-knit with updated data to refresh all outputs.*
