---
title: "Survival Data Concepts"
subtitle: "Survival Analysis — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(survival)
library(knitr)

policyholders <- read_csv("L2_policyholders.csv") %>%
  mutate(
    # Survival time = policy duration in months
    time  = policy_duration_months,
    # Event = lapse (1 = lapsed, 0 = still active = right-censored)
    event = if_else(policy_duration_months < 60, 1, 0)
  )
```

# Overview

This report introduces survival data structure and the `Surv()` object
using the **`r nrow(policyholders)`-policyholder** portfolio.

Each policyholder contributes one observation:

- **Survival time** (`time`): months the policy was in force
- **Event indicator** (`event`): 1 = policy lapsed (event occurred),
  0 = policy still active at observation end (right-censored)

Of the **`r nrow(policyholders)`** policyholders,
**`r sum(policyholders$event)`**
(**`r paste0(round(mean(policyholders$event)*100,1),"%")`**)
experienced the event (lapse), and
**`r sum(1 - policyholders$event)`**
(**`r paste0(round(mean(1-policyholders$event)*100,1),"%")`**)
are right-censored.

---

# Survival Data Structure

## Raw Data Preview

```{r raw-preview}
policyholders %>%
  select(policyholder_id, policy_type, time, event) %>%
  head(10) %>%
  kable(caption = "Table 1: Survival Data Structure (first 10 rows)")
```

## Event vs Censored Distribution

```{r event-dist}
policyholders %>%
  mutate(Status = if_else(event == 1, "Lapsed (Event)", "Active (Censored)")) %>%
  count(Status) %>%
  mutate(`%` = paste0(round(n / sum(n) * 100, 1), "%")) %>%
  kable(caption = "Table 2: Event vs Censored Count")
```

---

# Censoring Visualisation

```{r censoring-plot, echo=TRUE, fig.width=9, fig.height=5,fig.cap="Figure 1: Censoring Plot — Sample of 40 Policyholders"}
set.seed(42)

sample_ids <- policyholders %>%
  slice_sample(n = 40) %>%
  arrange(time) %>%
  mutate(rank = row_number())

ggplot(sample_ids) +
  geom_segment(aes(x = 0, xend = time,
                   y = rank, yend = rank,
                   color = factor(event)),
               linewidth = 0.7) +
  geom_point(aes(x = time, y = rank,
                 color  = factor(event),
                 shape  = factor(event)),
             size = 2.5) +
  scale_color_manual(values = c("0" = "#2c7bb6", "1" = "#d7191c"),
                     labels = c("0" = "Censored", "1" = "Lapsed")) +
  scale_shape_manual(values = c("0" = 1, "1" = 16),
                     labels = c("0" = "Censored", "1" = "Lapsed")) +
  labs(title  = "Censoring Plot — 40 Sampled Policyholders",
       subtitle = "Each line = one policyholder's observation window",
       x = "Months in Force", y = "Policyholder (ranked by duration)",
       color = "Status", shape = "Status") +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

---

# The Surv() Object

```{r surv-object, echo=TRUE}
surv_obj <- Surv(time  = policyholders$time,
                 event = policyholders$event)

# Preview first 20 entries
surv_obj[1:20]
```

The `+` symbol after a value indicates a censored observation —
the event had not yet occurred when the policyholder was last observed.

---

# Basic survfit()

```{r survfit, echo=TRUE}
fit <- survfit(Surv(time, event) ~ 1, data = policyholders)
summary(fit, times = c(12, 24, 36, 48, 60))
```

```{r survfit-plot, fig.width=8, fig.height=4,fig.cap="Figure 2: Kaplan-Meier Survival Curve (Overall Portfolio)"}
plot_data <- tibble(
  time     = fit$time,
  surv     = fit$surv,
  lower    = fit$lower,
  upper    = fit$upper
)

ggplot(plot_data, aes(x = time, y = surv)) +
  geom_step(color = "#2c7bb6", linewidth = 0.9) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              alpha = 0.15, fill = "#2c7bb6") +
  scale_y_continuous(limits = c(0, 1),
                     labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(0, 72, 12)) +
  labs(title    = "Kaplan-Meier Survival Curve — Overall Portfolio",
       subtitle = "Shaded band = 95% confidence interval",
       x = "Months in Force", y = "Survival Probability S(t)") +
  theme_minimal(base_size = 12)
```

---

# Survival by Policy Type

```{r survfit-by-type, echo=TRUE}
fit_type <- survfit(Surv(time, event) ~ policy_type,
                    data = policyholders)
```

```{r survfit-type-plot, fig.width=8, fig.height=5,fig.cap="Figure 3: Kaplan-Meier Curves by Policy Type"}
# Extract plot data for all strata
strata_names <- names(fit_type$strata)

plot_type <- tibble(
  time   = fit_type$time,
  surv   = fit_type$surv,
  strata = rep(strata_names,
               times = fit_type$strata)
) %>%
  mutate(strata = str_remove(strata, "policy_type="))

ggplot(plot_type, aes(x = time, y = surv, color = strata)) +
  geom_step(linewidth = 0.8) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(0, 1),
                     labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(0, 72, 12)) +
  labs(title  = "Kaplan-Meier Survival Curves by Policy Type",
       x = "Months in Force", y = "Survival Probability S(t)",
       color = "Policy Type") +
  theme_minimal(base_size = 12)
```

---

# Survival Rates at Key Durations

```{r survival-table}
times_of_interest <- c(12, 24, 36, 48)

map_dfr(times_of_interest, function(t) {
  summary(fit_type, times = t)$strata %>%
    tibble(policy_type = .,
           surv = summary(fit_type, times = t)$surv) %>%
    mutate(months = t,
           policy_type = str_remove(policy_type, "policy_type="),
           surv = paste0(round(surv * 100, 1), "%"))
}) %>%
  pivot_wider(names_from = months,
              values_from = surv,
              names_prefix = "Month ") %>%
  kable(caption = "Table 3: Survival Rate by Policy Type at Key Durations")
```

---

# Key Findings

- Survival analysis correctly accounts for **right-censoring** —
  policyholders still active at the observation end contribute
  information without being treated as events.
- The `Surv()` object is the foundation of all survival modelling in R.
  Every function in the survival package — `survfit()`, `survreg()`,
  `coxph()` — takes a `Surv()` object as its response.
- The overall portfolio survival curve shows the proportion of
  policyholders still active at each duration point.
- Survival rates differ meaningfully by product line — these differences
  are formally tested in **Lesson 14** using the log-rank test alongside
  the Kaplan-Meier and Nelson-Aalen estimators.

---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
