---
title: "Machine Learning Models for Churn: RF & XGBoost"
subtitle: "Applied Predictive & Causal Analytics — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(randomForest)
library(xgboost)
library(pROC)
library(knitr)

set.seed(42)

policyholders <- read_csv("L2_policyholders.csv") %>%
  mutate(
    churned     = if_else(policy_duration_months < 24, 1, 0),
    age_band    = cut(age, breaks = c(0, 35, 50, 65, 100),
                      labels = c("Under35", "Age35_50", "Age50_65", "Over65")),
    policy_type = factor(policy_type),
    region      = factor(region),
    gender      = factor(gender),
    claim_to_premium = claim_amount / annual_premium
  )

train_idx <- sample(nrow(policyholders), 0.7 * nrow(policyholders))
train <- policyholders[train_idx, ]
test  <- policyholders[-train_idx, ]
```

# Overview

This report fits two tree-based machine-learning models — **Random Forest**
and **XGBoost** — to predict churn, and benchmarks them against the
logistic regression GLM from Lesson 22.

Data: **`r nrow(policyholders)`** policyholders
(**`r nrow(train)`** train / **`r nrow(test)`** test).

---

# Model 1 — Random Forest

```{r rf-fit, echo=TRUE}
rf_model <- randomForest(
  factor(churned) ~ age_band + policy_type + annual_premium +
                    claim_to_premium + region + gender,
  data       = train,
  ntree      = 500,
  importance = TRUE
)
```

```{r rf-predict, echo=TRUE}
rf_prob <- predict(rf_model, newdata = test, type = "prob")[, "1"]
rf_auc  <- as.numeric(auc(roc(test$churned, rf_prob, quiet = TRUE)))
```

## Variable Importance

```{r rf-importance, fig.width=8, fig.height=4, fig.cap="Figure 1: Random Forest Variable Importance"}
importance(rf_model) %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  arrange(desc(MeanDecreaseGini)) %>%
  ggplot(aes(x = reorder(variable, MeanDecreaseGini),
             y = MeanDecreaseGini)) +
  geom_col(fill = "#2c7bb6", alpha = 0.85) +
  coord_flip() +
  labs(title = "Random Forest — Variable Importance",
       x = NULL, y = "Mean Decrease in Gini") +
  theme_minimal(base_size = 12)
```

---

# Model 2 — XGBoost

```{r xgb-prep, echo=TRUE}
# XGBoost needs a numeric matrix; one-hot encode predictors
feature_formula <- ~ age_band + policy_type + annual_premium +
                     claim_to_premium + region + gender - 1

train_matrix <- model.matrix(feature_formula, data = train)
test_matrix  <- model.matrix(feature_formula, data = test)

dtrain <- xgb.DMatrix(data = train_matrix, label = train$churned)
dtest  <- xgb.DMatrix(data = test_matrix,  label = test$churned)
```

```{r xgb-fit, echo=TRUE}
# NOTE on xgboost versions:
# In xgboost 2.x / 3.x the high-level xgboost() wrapper was rewritten with a
# scikit-learn-style interface: it now expects x = / y = (not data = / label =),
# learning_rate (not eta), and no longer accepts verbose. Calling the old
# xgboost(data = dtrain, eta = ..., verbose = 0) form therefore errors with
# "argument 'y' is missing".
#
# The low-level xgb.train() below is STABLE across all versions because it
# takes an xgb.DMatrix plus a params list. Prefer it for reproducible code.
params <- list(
  objective   = "binary:logistic",
  max_depth   = 4,
  eta         = 0.1,          # learning rate (xgb.train still accepts 'eta')
  eval_metric = "auc"
)

xgb_model <- xgb.train(
  params  = params,
  data    = dtrain,
  nrounds = 100,
  verbose = 0
)
```

```{r xgb-predict, echo=TRUE}
xgb_prob <- predict(xgb_model, newdata = dtest)
xgb_auc  <- as.numeric(auc(roc(test$churned, xgb_prob, quiet = TRUE)))
```

## Feature Importance

```{r xgb-importance, fig.width=8, fig.height=4, fig.cap="Figure 2: XGBoost Feature Importance (top 10)"}
xgb.importance(model = xgb_model) %>%
  head(10) %>%
  ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "#1a9641", alpha = 0.85) +
  coord_flip() +
  labs(title = "XGBoost — Feature Importance (Gain)",
       x = NULL, y = "Gain") +
  theme_minimal(base_size = 12)
```

---

# Benchmark: GLM vs RF vs XGBoost

```{r glm-benchmark, echo=TRUE}
# Refit GLM for comparison (Lesson 22)
glm_model <- glm(churned ~ age_band + policy_type + annual_premium +
                           claim_to_premium + region,
                 data = train, family = binomial)

glm_prob <- predict(glm_model, newdata = test, type = "response")
glm_auc  <- as.numeric(auc(roc(test$churned, glm_prob, quiet = TRUE)))
```

```{r benchmark-table}
tibble(
  Model = c("Logistic GLM", "Random Forest", "XGBoost"),
  AUC   = round(c(glm_auc, rf_auc, xgb_auc), 3),
  Interpretability = c("High (odds ratios)",
                       "Medium (importance only)",
                       "Medium (importance only)")
) %>%
  arrange(desc(AUC)) %>%
  kable(caption = "Table 1: Model Benchmark — AUC and Interpretability")
```

```{r best-model}
aucs <- c(`Logistic GLM` = glm_auc,
          `Random Forest` = rf_auc,
          XGBoost = xgb_auc)
best_model <- names(which.max(aucs))
```

The highest AUC is achieved by **`r best_model`**.

## ROC Curves Overlaid

```{r roc-overlay, fig.width=7, fig.height=5, fig.cap="Figure 3: ROC Curves — GLM vs RF vs XGBoost"}
roc_data <- bind_rows(
  tibble(fpr = 1 - roc(test$churned, glm_prob, quiet = TRUE)$specificities,
         tpr = roc(test$churned, glm_prob, quiet = TRUE)$sensitivities,
         model = "Logistic GLM"),
  tibble(fpr = 1 - roc(test$churned, rf_prob, quiet = TRUE)$specificities,
         tpr = roc(test$churned, rf_prob, quiet = TRUE)$sensitivities,
         model = "Random Forest"),
  tibble(fpr = 1 - roc(test$churned, xgb_prob, quiet = TRUE)$specificities,
         tpr = roc(test$churned, xgb_prob, quiet = TRUE)$sensitivities,
         model = "XGBoost")
)

ggplot(roc_data, aes(x = fpr, y = tpr, color = model)) +
  geom_line(linewidth = 0.9) +
  geom_abline(linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("Logistic GLM"  = "#d7191c",
                                 "Random Forest" = "#2c7bb6",
                                 "XGBoost"       = "#1a9641")) +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "ROC Curve Comparison",
       x = "False Positive Rate", y = "True Positive Rate",
       color = "Model") +
  theme_minimal(base_size = 12)
```

---

# Key Findings

```{r kf-vals, include=FALSE}
auc_gap <- round(max(aucs) - glm_auc, 3)
```

- **`r best_model`** achieves the highest test-set AUC. The gap over the
  interpretable GLM benchmark is **`r auc_gap`** AUC points.
- Tree-based models capture **non-linear effects and interactions**
  automatically, without manual feature engineering — but at the cost of
  the direct interpretability the GLM provides through odds ratios.
- **Variable importance** (Figures 1–2) offers partial interpretability:
  it ranks predictors but does not give signed, quantitative effects.
- The practical choice depends on context: regulated pricing favours the
  transparent GLM; an internal retention-targeting tool may favour the
  higher-AUC machine-learning model.
- In **Lesson 24**, we shift from prediction to **customer segmentation**
  using unsupervised methods (k-means and hierarchical clustering).

---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
