---
title: "Policyholder Data Cleaning Report"
subtitle: "Tidy Data & Advanced Cleaning — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(lubridate)
library(knitr)

raw <- read_csv("L3_policyholders.csv")
```

# Overview

This report documents the cleaning pipeline applied to the raw policyholder extract.
The raw dataset contains **`r nrow(raw)`** records across
**`r ncol(raw)`** fields. Before any analysis can begin, the data must be
standardised, parsed, and reshaped into a tidy format.

---

# Step 1 — Standardise Text Fields (stringr)

Raw `gender` and `policy_type` fields contain inconsistent capitalisation,
leading/trailing whitespace, and non-standard abbreviations.

```{r clean-text, echo=TRUE}
cleaned <- raw %>%
  mutate(
    gender = str_to_lower(str_trim(gender)),
    gender = case_when(
      gender %in% c("m", "male")   ~ "Male",
      gender %in% c("f", "female") ~ "Female",
      TRUE ~ NA_character_
    ),
    policy_type = str_to_title(str_trim(policy_type))
  )
```

```{r text-check}
cleaned %>%
  count(policy_type, gender) %>%
  kable(caption = "Table 1: Standardised Policy Type × Gender Counts")
```

---

# Step 2 — Parse Dates (lubridate)

The `policy_start_date` field was supplied in mixed formats across source systems.
`parse_date_time()` handles all observed formats in a single call.

```{r parse-dates, echo=TRUE}
cleaned <- cleaned %>%
  mutate(
    policy_start_date = parse_date_time(
      policy_start_date,
      orders = c("ymd", "dmy", "mdy", "d b Y")
    ),
    policy_year  = year(policy_start_date),
    policy_month = month(policy_start_date, label = TRUE)
  )
```

```{r date-check}
cleaned %>%
  count(policy_year) %>%
  kable(caption = "Table 2: Policy Count by Inception Year")
```

---

# Step 3 — Reshape with tidyr

To compare premium bands across policy types, we create a wide summary table
using `pivot_wider()`.

```{r reshape, echo=TRUE}
premium_summary <- cleaned %>%
  filter(!is.na(annual_premium)) %>%
  mutate(premium_band = case_when(
    annual_premium < 1000 ~ "Low",
    annual_premium < 2000 ~ "Mid",
    TRUE                  ~ "High"
  )) %>%
  count(policy_type, premium_band) %>%
  pivot_wider(names_from = premium_band, values_from = n, values_fill = 0)
```

```{r reshape-table}
premium_summary %>%
  kable(caption = "Table 3: Policyholder Count by Policy Type and Premium Band")
```

---

# Step 4 — Handle Missing Values

```{r missing, echo=TRUE}
cleaned <- cleaned %>%
  mutate(
    age            = suppressWarnings(as.numeric(age)),
    annual_premium = as.numeric(annual_premium)
  ) %>%
  filter(!is.na(age), !is.na(annual_premium))
```

```{r missing-summary}
tibble(
  Field           = c("age", "annual_premium", "gender", "policy_start_date"),
  `Missing (raw)` = c(
    sum(is.na(suppressWarnings(as.numeric(raw$age)))),
    sum(is.na(as.numeric(raw$annual_premium))),
    sum(is.na(raw$gender)),
    sum(is.na(raw$policy_start_date))
  ),
  `Missing (cleaned)` = c(
    sum(is.na(cleaned$age)),
    sum(is.na(cleaned$annual_premium)),
    sum(is.na(cleaned$gender)),
    sum(is.na(cleaned$policy_start_date))
  )
) %>%
  kable(caption = "Table 4: Missing Value Summary Before and After Cleaning")
```

---

# Clean Dataset — Final Preview

```{r final-preview}
cleaned %>%
  select(policyholder_id, age, gender, policy_type,
         annual_premium, policy_start_date, policy_year) %>%
  head(8) %>%
  kable(caption = "Table 5: Cleaned Dataset Preview (first 8 rows)")
```

**`r nrow(cleaned)`** records retained after cleaning
(**`r nrow(raw) - nrow(cleaned)`** removed due to missing critical fields).

---

*Report generated automatically using R Markdown. Data source: `policyholders_messy.csv`.*
