---
title: "Churn Curve Smoothing & Graduation"
subtitle: "Churn & Retention Modelling — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(knitr)

policyholders <- read_csv("L2_policyholders.csv") %>%
  mutate(
    time  = policy_duration_months,
    event = if_else(policy_duration_months < 60, 1, 0)
  )
```

# Overview

This report demonstrates **graduation** — the actuarial practice of
smoothing crude (observed) churn rates into a smooth, usable curve while
preserving the underlying pattern in the data.

We compute crude monthly churn hazards, then apply three smoothing
methods — LOESS, smoothing spline, and a moving average — and assess the
trade-off between **smoothness** and **adherence** to the raw data.

Portfolio: **`r nrow(policyholders)`** policyholders,
**`r sum(policyholders$event)`** churn events.

---

# Crude Churn Rates

## Compute Raw Monthly Hazard

```{r crude-rates, echo=TRUE}
crude <- policyholders %>%
  group_by(tenure = time) %>%
  summarise(
    at_risk = n(),
    events  = sum(event),
    .groups = "drop"
  ) %>%
  arrange(tenure) %>%
  mutate(
    # Number still at risk = total who reached this tenure
    n_reaching = sum(at_risk) - lag(cumsum(at_risk), default = 0),
    crude_hazard = events / n_reaching
  ) %>%
  filter(n_reaching > 0, tenure <= 60)
```

```{r crude-table}
crude %>%
  select(tenure, n_reaching, events, crude_hazard) %>%
  filter(tenure %in% c(6, 12, 18, 24, 30, 36, 42, 48)) %>%
  mutate(crude_hazard = round(crude_hazard, 4)) %>%
  rename(`Tenure` = tenure, `At Risk` = n_reaching,
         `Events` = events, `Crude Hazard` = crude_hazard) %>%
  kable(caption = "Table 1: Crude Monthly Churn Hazard (sampled tenures)")
```

The crude hazards are noisy — especially at longer tenures where the
at-risk population is small. Graduation addresses this.

---

# Method 1 — LOESS Smoothing

```{r loess, echo=TRUE}
loess_fit <- loess(crude_hazard ~ tenure,
                   data = crude,
                   span = 0.5)

crude$loess_smooth <- predict(loess_fit)
```

---

# Method 2 — Smoothing Spline

```{r spline, echo=TRUE}
spline_fit <- smooth.spline(x = crude$tenure,
                            y = crude$crude_hazard,
                            spar = 0.6)

crude$spline_smooth <- predict(spline_fit, crude$tenure)$y
```

---

# Method 3 — Moving Average

```{r moving-avg, echo=TRUE}
crude <- crude %>%
  arrange(tenure) %>%
  mutate(
    ma_smooth = zoo::rollmean(crude_hazard, k = 5,
                              fill = NA, align = "center")
  )
```

---

# Visual Comparison

```{r comparison-plot, fig.width=9, fig.height=5,fig.cap="Figure 1: Crude vs Graduated Churn Hazards"}
crude %>%
  select(tenure, crude_hazard, loess_smooth,
         spline_smooth, ma_smooth) %>%
  pivot_longer(-tenure, names_to = "method", values_to = "hazard") %>%
  mutate(method = recode(method,
                         crude_hazard  = "Crude (raw)",
                         loess_smooth  = "LOESS",
                         spline_smooth = "Smoothing Spline",
                         ma_smooth     = "Moving Average")) %>%
  ggplot(aes(x = tenure, y = hazard, color = method)) +
  geom_line(data = . %>% filter(method == "Crude (raw)"),
            linewidth = 0.5, alpha = 0.5) +
  geom_point(data = . %>% filter(method == "Crude (raw)"),
             size = 1, alpha = 0.5) +
  geom_line(data = . %>% filter(method != "Crude (raw)"),
            linewidth = 1) +
  scale_color_manual(values = c("Crude (raw)"      = "grey50",
                                 "LOESS"            = "#d7191c",
                                 "Smoothing Spline" = "#2c7bb6",
                                 "Moving Average"   = "#1a9641")) +
  labs(title  = "Crude vs Graduated Churn Hazards",
       x = "Tenure (Months)", y = "Monthly Churn Hazard",
       color = "Method") +
  theme_minimal(base_size = 12)
```

---

# Smoothness vs Adherence

```{r smoothness-adherence}
# Adherence: sum of squared differences from crude
# Smoothness: sum of squared second differences

assess <- function(smoothed, crude_vals) {
  valid <- !is.na(smoothed) & !is.na(crude_vals)
  s <- smoothed[valid]
  c <- crude_vals[valid]
  adherence  <- sum((s - c)^2)
  smoothness <- sum(diff(diff(s))^2)
  c(adherence = adherence, smoothness = smoothness)
}

results <- bind_rows(
  LOESS              = assess(crude$loess_smooth,  crude$crude_hazard),
  `Smoothing Spline` = assess(crude$spline_smooth, crude$crude_hazard),
  `Moving Average`   = assess(crude$ma_smooth,     crude$crude_hazard),
  .id = "Method"
)

results %>%
  mutate(adherence  = round(adherence, 6),
         smoothness = round(smoothness, 8)) %>%
  rename(`Adherence (SS deviation)` = adherence,
         `Roughness (SS 2nd diff)`  = smoothness) %>%
  kable(caption = "Table 2: Smoothness vs Adherence Trade-off")
```

Lower **adherence** values mean the smoothed curve stays closer to the
raw data. Lower **roughness** values mean the curve is smoother. The two
objectives compete — the actuary's task is to balance them.

---

# Graduated Retention Curve

```{r graduated-survival, fig.width=8, fig.height=5,fig.cap="Figure 2: Retention Curve from Graduated Hazards (LOESS)"}
crude_surv <- crude %>%
  arrange(tenure) %>%
  mutate(
    haz_graduated = pmax(loess_smooth, 0),
    surv_graduated = cumprod(1 - haz_graduated)
  )

ggplot(crude_surv, aes(x = tenure, y = surv_graduated)) +
  geom_line(color = "#2c7bb6", linewidth = 0.9) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(0, 60, 12)) +
  labs(title    = "Graduated Retention Curve",
       subtitle = "Derived from LOESS-smoothed monthly hazards",
       x = "Tenure (Months)", y = "Retention S(t)") +
  theme_minimal(base_size = 12)
```

---

# Key Findings

- **Crude churn hazards** are volatile, especially at longer tenures
  where the at-risk population shrinks. They are unsuitable for direct
  use in pricing or reserving.
- **Graduation** smooths these rates while preserving the underlying
  pattern. LOESS and smoothing splines both produce continuous curves;
  the moving average is simpler but loses data at the series ends.
- The **smoothness-adherence trade-off** (Table 2) is the central tension
  in graduation: too much smoothing distorts the real pattern, too little
  leaves noise. The actuary selects the balance appropriate to the use.
- The **graduated retention curve** (Figure 2) is built by accumulating
  the smoothed hazards via `cumprod(1 - h)` — the discrete survival
  relationship.
- In **Lesson 20**, we bring the full Part VI toolkit together in an
  applied end-to-end churn analysis project.

---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
