---
title: "Parametric Churn Models: Weibull & Log-Logistic"
subtitle: "Churn & Retention Modelling — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(survival)
library(flexsurv)
library(knitr)

policyholders <- read_csv("L2_policyholders.csv") %>%
  mutate(
    time  = policy_duration_months,
    event = if_else(policy_duration_months < 60, 1, 0)
  )

fit_km <- survfit(Surv(time, event) ~ 1, data = policyholders)
```

# Overview

This report fits and compares two parametric churn models — **Weibull**
and **Log-Logistic** — to the policyholder retention data, focusing on
the difference in their hazard shapes and which better describes churn
behaviour.

Portfolio: **`r nrow(policyholders)`** policyholders,
**`r sum(policyholders$event)`** churn events,
**`r sum(1 - policyholders$event)`** censored.

---

# Why Log-Logistic for Churn?

The Weibull hazard is **monotone** — it can only increase or decrease
with tenure. The Log-Logistic hazard can be **non-monotone** —
rising then falling — which often matches real churn: low risk at
onboarding, a peak in the early-mid tenure, then declining risk for
loyal long-tenure customers.

---

# Fit Both Models

```{r fit-models, echo=TRUE}
fit_weib <- survreg(Surv(time, event) ~ 1,
                    data = policyholders, dist = "weibull")

fit_llog <- survreg(Surv(time, event) ~ 1,
                    data = policyholders, dist = "loglogistic")
```

## Parameter Extraction

```{r extract-params, echo=TRUE}
# Weibull
k_weib      <- 1 / fit_weib$scale
lambda_weib <- exp(coef(fit_weib)[1])

# Log-logistic
shape_llog  <- 1 / fit_llog$scale
scale_llog  <- exp(coef(fit_llog)[1])

cat("Weibull       — shape:", round(k_weib, 3),
    " scale:", round(lambda_weib, 1), "\n")
cat("Log-Logistic  — shape:", round(shape_llog, 3),
    " scale:", round(scale_llog, 1), "\n")
```

---

# Survival Curves vs Kaplan-Meier

```{r survival-compare, fig.width=8, fig.height=5,fig.cap="Figure 1: KM vs Weibull vs Log-Logistic Survival Curves"}
t_seq <- seq(0, 72, by = 0.5)

S_weib <- exp(-(t_seq / lambda_weib)^k_weib)
S_llog <- 1 / (1 + (t_seq / scale_llog)^shape_llog)

km_df <- tibble(time = c(0, fit_km$time),
                surv = c(1, fit_km$surv),
                model = "Kaplan-Meier")

parametric_df <- bind_rows(
  tibble(time = t_seq, surv = S_weib, model = "Weibull"),
  tibble(time = t_seq, surv = S_llog, model = "Log-Logistic")
)

ggplot() +
  geom_step(data = km_df, aes(x = time, y = surv, color = model),
            linewidth = 0.8) +
  geom_line(data = parametric_df,
            aes(x = time, y = surv, color = model),
            linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Kaplan-Meier" = "#2c7bb6",
                                 "Weibull"      = "#d7191c",
                                 "Log-Logistic" = "#1a9641")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(0, 72, 12)) +
  labs(title    = "Survival Curves: KM vs Parametric Churn Models",
       subtitle = "Step = KM | Dashed = parametric fits",
       x = "Tenure (Months)", y = "Survival Probability S(t)",
       color = "Model") +
  theme_minimal(base_size = 12)
```

---

# Hazard Functions — The Key Difference

```{r hazard-compare, fig.width=8, fig.height=4,fig.cap="Figure 2: Hazard Functions — Weibull (monotone) vs Log-Logistic (non-monotone)"}
h_weib <- (k_weib / lambda_weib) * (t_seq / lambda_weib)^(k_weib - 1)

# Log-logistic hazard
num    <- (shape_llog / scale_llog) * (t_seq / scale_llog)^(shape_llog - 1)
den    <- 1 + (t_seq / scale_llog)^shape_llog
h_llog <- num / den

tibble(
  time   = rep(t_seq, 2),
  hazard = c(h_weib, h_llog),
  model  = rep(c("Weibull", "Log-Logistic"), each = length(t_seq))
) %>%
  filter(time > 0) %>%
  ggplot(aes(x = time, y = hazard, color = model)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("Weibull"      = "#d7191c",
                                 "Log-Logistic" = "#1a9641")) +
  labs(title    = "Hazard Functions: Weibull vs Log-Logistic",
       subtitle = "Weibull is monotone; Log-Logistic can rise then fall",
       x = "Tenure (Months)", y = "Churn Hazard h(t)",
       color = "Model") +
  theme_minimal(base_size = 12)
```

---

# Model Comparison

```{r aic-table}
tibble(
  Model = c("Weibull", "Log-Logistic"),
  `Log-Likelihood` = round(c(fit_weib$loglik[2],
                              fit_llog$loglik[2]), 2),
  AIC = round(c(AIC(fit_weib), AIC(fit_llog)), 2),
  BIC = round(c(BIC(fit_weib), BIC(fit_llog)), 2)
) %>%
  arrange(AIC) %>%
  mutate(`ΔAIC` = round(AIC - min(AIC), 2)) %>%
  kable(caption = "Table 1: Weibull vs Log-Logistic — Model Comparison")
```

```{r best}
best_churn <- if (AIC(fit_llog) < AIC(fit_weib)) "Log-Logistic" else "Weibull"
```

The **`r best_churn`** model is preferred by AIC for this churn data.

---

# Retention Estimates at Key Tenures

```{r retention-table}
tenures <- c(12, 24, 36, 48, 60)

tibble(
  `Tenure (Months)` = tenures,
  `Kaplan-Meier` = paste0(
    round(summary(fit_km, times = tenures)$surv * 100, 1), "%"),
  `Weibull` = paste0(
    round(exp(-(tenures / lambda_weib)^k_weib) * 100, 1), "%"),
  `Log-Logistic` = paste0(
    round((1 / (1 + (tenures / scale_llog)^shape_llog)) * 100, 1), "%")
) %>%
  kable(caption = "Table 2: Retention Rate Estimates at Key Tenures")
```

---

# Key Findings

```{r kf-vals, include=FALSE}
delta <- round(abs(AIC(fit_weib) - AIC(fit_llog)), 2)
```

- The **Weibull** hazard is monotone (shape k =
  **`r round(k_weib, 3)`**), implying churn risk changes in a single
  direction with tenure.
- The **Log-Logistic** hazard is non-monotone when its shape parameter
  exceeds 1 (shape = **`r round(shape_llog, 3)`**) — capturing an
  early-tenure peak in churn followed by declining risk, a common
  retention pattern.
- The **`r best_churn`** model achieves the lower AIC
  (ΔAIC = **`r delta`**) and is preferred for this portfolio.
- The choice of distribution materially affects long-tenure retention
  projections — visible in the divergence between curves beyond 48
  months in Figure 1.
- In **Lesson 19**, we apply **smoothing and graduation** techniques
  to produce regulatory-grade churn curves from these parametric fits.

---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
