---
title: "Applied Churn Analysis Project"
subtitle: "Churn & Retention Modelling — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(survival)
library(knitr)

policyholders <- read_csv("L2_policyholders.csv") %>%
  mutate(
    time     = policy_duration_months,
    event    = if_else(policy_duration_months < 60, 1, 0),
    age_band = cut(age, breaks = c(0, 35, 50, 65, 100),
                   labels = c("Under 35", "35-50", "50-65", "Over 65"))
  )
```

# Executive Summary

This end-to-end churn analysis applies the full Part VI toolkit to the
**`r nrow(policyholders)`-policyholder** portfolio: data structuring
(Lesson 17), parametric modelling (Lesson 18), and graduation (Lesson 19).

Headline metrics:

- Overall churn rate (tenure < 60 months):
  **`r paste0(round(mean(policyholders$event)*100, 1), "%")`**
- Total churn events: **`r sum(policyholders$event)`**
- Highest-risk segment is identified and a retention recommendation made.

---

# Step 1 — Segment-Level Churn Profile

```{r segment-profile}
segment_summary <- policyholders %>%
  group_by(policy_type) %>%
  summarise(
    Customers   = n(),
    Churned     = sum(event),
    `Churn Rate` = mean(event),
    `Median Tenure` = median(time),
    .groups = "drop"
  ) %>%
  arrange(desc(`Churn Rate`))
```

```{r segment-table}
segment_summary %>%
  mutate(`Churn Rate` = paste0(round(`Churn Rate` * 100, 1), "%")) %>%
  kable(caption = "Table 1: Churn Profile by Policy Type (highest risk first)")
```

```{r highest-risk}
highest_risk <- segment_summary$policy_type[1]
highest_rate <- segment_summary$`Churn Rate`[1]
```

The highest-risk segment is **`r highest_risk`**
(churn rate **`r paste0(round(highest_rate*100,1),"%")`**).

---

# Step 2 — Fit Survival Model by Segment

```{r fit-by-segment, echo=TRUE}
fit_segments <- survfit(Surv(time, event) ~ policy_type,
                        data = policyholders)
```

```{r segment-survival-plot, fig.width=8, fig.height=5,fig.cap="Figure 1: Kaplan-Meier Retention by Policy Type"}
strata_names <- names(fit_segments$strata)

km_data <- tibble(
  time   = fit_segments$time,
  surv   = fit_segments$surv,
  strata = rep(strata_names, times = fit_segments$strata)
) %>%
  mutate(policy_type = str_remove(strata, "policy_type="))

ggplot(km_data, aes(x = time, y = surv, color = policy_type)) +
  geom_step(linewidth = 0.8) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(0, 72, 12)) +
  labs(title = "Retention by Policy Type",
       x = "Tenure (Months)", y = "Retention S(t)",
       color = "Policy Type") +
  theme_minimal(base_size = 12)
```

---

# Step 3 — Parametric Model & Median Lifetime

```{r parametric-by-segment, echo=TRUE}
median_lifetimes <- policyholders %>%
  group_by(policy_type) %>%
  group_modify(~ {
    fit <- survreg(Surv(time, event) ~ 1,
                   data = .x, dist = "weibull")
    median_t <- predict(fit, newdata = data.frame(1),
                        type = "quantile", p = 0.5)[1]
    tibble(median_lifetime = median_t)
  }) %>%
  ungroup()
```

```{r lifetime-table}
median_lifetimes %>%
  arrange(median_lifetime) %>%
  mutate(median_lifetime = round(median_lifetime, 1)) %>%
  rename(`Policy Type` = policy_type,
         `Median Lifetime (Months)` = median_lifetime) %>%
  kable(caption = "Table 2: Weibull Median Customer Lifetime by Segment")
```

---

# Step 4 — Customer Lifetime Value Impact

```{r clv, echo=TRUE}
clv_analysis <- policyholders %>%
  group_by(policy_type) %>%
  summarise(
    avg_premium    = mean(annual_premium),
    median_tenure  = median(time),
    .groups = "drop"
  ) %>%
  left_join(median_lifetimes, by = "policy_type") %>%
  mutate(
    # Simple CLV proxy: monthly premium × expected lifetime
    monthly_premium = avg_premium / 12,
    estimated_clv   = monthly_premium * median_lifetime
  )
```

```{r clv-table}
clv_analysis %>%
  mutate(
    avg_premium     = paste0("£", round(avg_premium, 0)),
    median_lifetime = round(median_lifetime, 1),
    estimated_clv   = paste0("£", format(round(estimated_clv, 0),
                                          big.mark = ","))
  ) %>%
  select(`Policy Type` = policy_type,
         `Avg Annual Premium` = avg_premium,
         `Median Lifetime (Months)` = median_lifetime,
         `Estimated CLV` = estimated_clv) %>%
  kable(caption = "Table 3: Estimated Customer Lifetime Value by Segment")
```

```{r clv-plot, fig.width=8, fig.height=4,fig.cap="Figure 2: Estimated CLV by Policy Type"}
clv_analysis %>%
  ggplot(aes(x = reorder(policy_type, estimated_clv),
             y = estimated_clv, fill = policy_type)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = paste0("£", format(round(estimated_clv, 0),
                                            big.mark = ","))),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£"),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Estimated Customer Lifetime Value by Segment",
       x = NULL, y = "Estimated CLV (£)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

# Step 5 — Retention Recommendation

```{r recommendation, echo=TRUE}
# Priority score: high churn × high CLV = high retention priority
priority <- segment_summary %>%
  left_join(clv_analysis %>% select(policy_type, estimated_clv),
            by = "policy_type") %>%
  mutate(
    churn_rate     = Churned / Customers,
    priority_score = churn_rate * estimated_clv
  ) %>%
  arrange(desc(priority_score))
```

```{r priority-table}
priority %>%
  mutate(
    churn_rate     = paste0(round(churn_rate * 100, 1), "%"),
    estimated_clv  = paste0("£", format(round(estimated_clv, 0),
                                         big.mark = ",")),
    priority_score = round(priority_score, 0)
  ) %>%
  select(`Policy Type` = policy_type,
         `Churn Rate` = churn_rate,
         `Est. CLV` = estimated_clv,
         `Priority Score` = priority_score) %>%
  kable(caption = "Table 4: Retention Priority Ranking (churn × CLV)")
```

```{r top-priority}
top_segment <- priority$policy_type[1]
```

**Recommendation:** Focus retention resources on the **`r top_segment`**
segment, which combines elevated churn risk with high customer lifetime
value — yielding the greatest expected return on retention investment.

---

# Key Findings

- The **`r highest_risk`** segment has the highest raw churn rate, but
  retention priority should weight churn risk by customer value.
- On a **churn × CLV** priority basis, the **`r top_segment`** segment
  offers the greatest expected return on retention spend.
- The analysis integrates the full Part VI workflow: data structuring,
  Kaplan-Meier and Weibull modelling, and value-based prioritisation.
- This completes **Part VI — Churn & Retention Modelling**. In
  **Part VII**, we move to predictive and causal analytics — beginning
  with feature engineering for churn in Lesson 21.

---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
