---
title: "ACF, PACF & Decomposition of Premium Income"
subtitle: "Time Series Structure Analysis — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(forecast)
library(knitr)

monthly_premiums <- read_csv("L5_monthly_premiums.csv") %>%
  mutate(month = as.Date(month))

premium_ts <- ts(
  monthly_premiums$total_premium,
  start     = c(2019, 1),
  frequency = 12
)
```

# Overview

This report analyses the internal structure of the monthly premium income
series built in Lesson 6. The series spans
**`r format(min(monthly_premiums$month), '%B %Y')`** to
**`r format(max(monthly_premiums$month), '%B %Y')`**
(**`r length(premium_ts)`** observations, monthly frequency).

We apply three diagnostic tools: ACF, PACF, and classical decomposition.
The findings inform the ARIMA model specification in Lesson 8.

---

# Autocorrelation Analysis

## ACF — Autocorrelation Function

```{r acf-plot, echo=TRUE, fig.width=8, fig.height=4,fig.cap="Figure 1: ACF of Monthly Premium Income"}
ggAcf(premium_ts, lag.max = 36) +
  labs(title = "ACF — Monthly Premium Income",
       x = "Lag (Months)") +
  theme_minimal(base_size = 12)
```

## PACF — Partial Autocorrelation Function

```{r pacf-plot, echo=TRUE, fig.width=8, fig.height=4,fig.cap="Figure 2: PACF of Monthly Premium Income"}
ggPacf(premium_ts, lag.max = 36) +
  labs(title = "PACF — Monthly Premium Income",
       x = "Lag (Months)") +
  theme_minimal(base_size = 12)
```

## Reading the Plots

```{r acf-lag1, echo=TRUE}
# Extract lag-1 autocorrelation value
acf_vals <- acf(premium_ts, lag.max = 36, plot = FALSE)
cat("Lag-1 autocorrelation:", round(acf_vals$acf[2], 3), "\n")
cat("Lag-12 autocorrelation:", round(acf_vals$acf[13], 3), "\n")
```

The blue dashed lines represent the 95% confidence bounds
(±1.96/√n). Bars exceeding these bounds indicate statistically
significant autocorrelation at that lag.

---

# Classical Decomposition

## Additive Decomposition

```{r decompose, echo=TRUE}
decomp <- decompose(premium_ts, type = "additive")
```

```{r decompose-plot, echo=FALSE, fig.width=8, fig.height=6,fig.cap="Figure 3: Additive Decomposition of Monthly Premium Income"}
autoplot(decomp) +
  labs(title = "Additive Decomposition — Monthly Premium Income") +
  theme_minimal(base_size = 11)
```

## Seasonal Component

```{r seasonal-table}
seasonal_vals <- decomp$seasonal[1:12]
tibble(
  Month         = month.abb,
  `Seasonal Effect (£)` = round(seasonal_vals, 0)
) %>%
  kable(caption = "Table 1: Estimated Monthly Seasonal Effects")
```

## Trend Component

```{r trend-plot, echo=FALSE, fig.width=8, fig.height=3,fig.cap="Figure 4: Extracted Trend Component"}
tibble(
  month = monthly_premiums$month,
  trend = as.numeric(decomp$trend)
) %>%
  filter(!is.na(trend)) %>%
  ggplot(aes(x = month, y = trend)) +
  geom_line(color = "#2c7bb6", linewidth = 0.9) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  labs(title = "Trend Component",
       x = NULL, y = "Premium (£)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

# Implications for Modelling

```{r random-component, echo=FALSE, fig.width=8, fig.height=3,fig.cap="Figure 5: Random (Residual) Component"}
tibble(
  month  = monthly_premiums$month,
  random = as.numeric(decomp$random)
) %>%
  filter(!is.na(random)) %>%
  ggplot(aes(x = month, y = random)) +
  geom_line(color = "#d7191c", linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  labs(title = "Random Component",
       x = NULL, y = "Residual (£)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

The random component should behave like white noise — no obvious
pattern or trend. Remaining structure here would suggest the additive
model is insufficient and a multiplicative decomposition or
differencing may be required before ARIMA modelling.

---

# Key Findings

- The **ACF** shows slow decay, consistent with a trending series.
  Significant spikes at lags 12 and 24 confirm annual seasonality.
- The **PACF** cuts off sharply after lag 1, suggesting an AR(1)
  component may be sufficient once trend and seasonality are removed.
- **Decomposition** separates the series into trend, seasonal, and
  random components. The seasonal effect peaks in winter months
  (November–February) and dips in mid-year.
- The random component appears approximately centred on zero with no
  obvious remaining pattern — the additive model is a reasonable fit.
- These diagnostics directly inform the ARIMA order selection in
  **Lesson 8**.

---

*Report generated automatically using R Markdown.
Data source: `monthly_premiums.csv`.*
