---
title: "Survival Model Comparison & Reporting"
subtitle: "Survival Analysis — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(survival)
library(broom)
library(knitr)

policyholders <- read_csv("L2_policyholders.csv") %>%
  mutate(
    time  = policy_duration_months,
    event = if_else(policy_duration_months < 60, 1, 0),
    age_band = cut(age,
                   breaks = c(0, 35, 50, 65, 100),
                   labels = c("Under 35", "35-50", "50-65", "Over 65"))
  )
```

# Overview

This report compares a family of parametric survival models — with and
without covariates — and produces a formal model selection summary
suitable for actuarial documentation.

Portfolio: **`r nrow(policyholders)`** policyholders,
**`r sum(policyholders$event)`** events,
**`r sum(1 - policyholders$event)`** censored.

Candidate models combine two distributions (Exponential, Weibull)
with two specifications (intercept-only, covariate-adjusted).

---

# Fit Candidate Models

```{r fit-models, echo=TRUE}
models <- list(
  exp_null    = survreg(Surv(time, event) ~ 1,
                        data = policyholders, dist = "exponential"),
  weib_null   = survreg(Surv(time, event) ~ 1,
                        data = policyholders, dist = "weibull"),
  exp_cov     = survreg(Surv(time, event) ~ policy_type + age_band,
                        data = policyholders, dist = "exponential"),
  weib_cov    = survreg(Surv(time, event) ~ policy_type + age_band,
                        data = policyholders, dist = "weibull")
)
```

---

# Model Comparison Table

```{r comparison-table, echo=TRUE}
comparison <- map_dfr(names(models), function(nm) {
  m  <- models[[nm]]
  gl <- glance(m)
  tibble(
    Model        = nm,
    Distribution = m$dist,
    `Parameters` = length(coef(m)) + 1,
    `Log-Lik`    = round(as.numeric(logLik(m)), 1),
    AIC          = round(AIC(m), 1),
    BIC          = round(BIC(m), 1)
  )
}) %>%
  arrange(AIC) %>%
  mutate(`ΔAIC` = round(AIC - min(AIC), 1))
```

```{r comparison-kable}
comparison %>%
  kable(caption = "Table 1: Candidate Model Comparison (sorted by AIC)")
```

```{r best-model}
best_name  <- comparison$Model[1]
best_model <- models[[best_name]]

# Best covariate-adjusted model — used specifically to illustrate how
# policy type and age band affect lapse risk. This may differ from the
# globally AIC-best model above, since AIC penalises the extra
# covariate parameters and can favour the simpler intercept-only model
# even when the covariate effects are informative for segmentation.
cov_comparison <- comparison %>% filter(Model %in% c("exp_cov", "weib_cov"))
cov_name       <- cov_comparison$Model[1]
cov_model      <- models[[cov_name]]

uses_cov_model <- !grepl("null", best_name)
delta_cov_vs_best <- round(cov_comparison$AIC[1] - comparison$AIC[1], 1)

# Pre-build the explanatory sentence here (instead of inline) so the
# narrative text below stays short and readable.
model_note <- if (!uses_cov_model) {
  paste0("This is an intercept-only model — it does not include ",
         "`policy_type` or `age_band`, so it cannot, by itself, show ",
         "how lapse risk differs across segments. For that purpose, ",
         "the remainder of this report uses the best-fitting ",
         "**covariate-adjusted** model, **", cov_name, "** (ΔAIC = ",
         delta_cov_vs_best, " versus the global best) — chosen ",
         "specifically to illustrate covariate effects, which is the ",
         "focus of this lesson.")
} else {
  paste0("It already includes `policy_type` and `age_band`, so it is ",
         "used directly for the covariate analysis below.")
}
```

The model preferred overall by AIC is **`r best_name`**
(`r best_model$dist` distribution). `r model_note`

---

# Best Model: Coefficient Summary

```{r tidy-coefs, echo=TRUE}
tidy(cov_model) %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  kable(caption = paste0("Table 2: Coefficient Estimates — ",
                         cov_name, " (Covariate Model)"))
```

## Interpreting Covariate Effects

```{r hazard-ratios}
# exp(-coef/scale) gives hazard ratios for survreg AFT models.
# This formula is valid for both Weibull and Exponential fits, since
# Exponential is the special case of Weibull with scale = 1.
# We use cov_model (not best_model) here, because hazard ratios are
# only meaningful for a model that actually includes covariates.
coefs <- coef(cov_model)[-1]   # exclude intercept
scale <- cov_model$scale
hr    <- exp(-coefs / scale)

tibble(
  Covariate      = names(coefs),
  `Coefficient`  = round(coefs, 4),
  `Hazard Ratio` = round(hr, 3),
  Effect         = ifelse(hr > 1,
                          "Higher lapse risk",
                          "Lower lapse risk")
) %>%
  kable(caption = paste0("Table 3: Hazard Ratios — Covariate Effects (",
                         cov_model$dist, " model)"))
```

---

# Predicted Survival by Policy Type

```{r predicted-survival, fig.width=8, fig.height=5, fig.cap="Figure 1: Predicted Survival Curves by Policy Type (Covariate Model)"}
t_seq <- seq(0, 72, by = 1)

newdata <- expand_grid(
  policy_type = unique(policyholders$policy_type),
  age_band    = "35-50"
) %>%
  mutate(
    # age_band must carry the SAME factor levels as the training data,
    # otherwise predict() cannot match the fitted coefficients.
    age_band = factor(age_band, levels = levels(policyholders$age_band))
  )

pred_df <- map_dfr(1:nrow(newdata), function(i) {
  # Predicted quantiles → survival function
  # predict(..., type="quantile") always returns a matrix, even when
  # p has length > 1 with newdata having 1 row — flatten with as.numeric()
  p_seq <- seq(0.01, 0.99, by = 0.01)
  q     <- as.numeric(predict(cov_model,
                              newdata = newdata[i, ],
                              type    = "quantile",
                              p       = p_seq))
  tibble(
    policy_type = newdata$policy_type[i],
    time        = q,
    surv        = 1 - p_seq
  )
})

ggplot(pred_df, aes(x = time, y = surv, color = policy_type)) +
  geom_line(linewidth = 0.9) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(0, 1),
                     labels = scales::percent_format()) +
  scale_x_continuous(limits = c(0, 72), breaks = seq(0, 72, 12)) +
  labs(title    = "Predicted Survival by Policy Type",
       subtitle = paste0("Covariate model: ", cov_name,
                         " (age band 35-50)"),
       x = "Months in Force", y = "Survival Probability S(t)",
       color = "Policy Type") +
  theme_minimal(base_size = 12)
```

---

# Predicted Median Survival Time

```{r median-survival-table}
median_pred <- as.numeric(predict(cov_model,
                                  newdata = newdata,
                                  type    = "quantile",
                                  p       = 0.5))

newdata %>%
  mutate(`Median Survival (Months)` = round(median_pred, 1)) %>%
  select(`Policy Type` = policy_type,
         `Age Band` = age_band,
         `Median Survival (Months)`) %>%
  kable(caption = paste0("Table 4: Predicted Median Survival Time by Policy Type (",
                         cov_name, ")"))
```

---

# Key Findings

```{r kf-vals, include=FALSE}
# delta_cov_vs_best was already computed in the best-model chunk above;
# just build the explanatory sentence here for readability in the bullet.
aic_fit_note <- if (!uses_cov_model) {
  paste0("Since this is the intercept-only model, AIC indicates that ",
         "`policy_type` and `age_band` do not materially improve the ",
         "*overall* fit once the extra parameters are penalised.")
} else {
  paste0("This model already includes the covariates, so no separate ",
         "covariate model was needed for the analysis below.")
}
```

- Across all four candidate models, **`r best_name`**
  (`r best_model$dist` distribution) achieves the lowest AIC overall.
  `r aic_fit_note`
- For segmentation purposes, this report uses the best-fitting
  **covariate-adjusted** model, **`r cov_name`**
  (ΔAIC = **`r delta_cov_vs_best`** versus the global best).
  Even when a simpler model wins narrowly on AIC, the covariate model
  remains useful for illustrating *relative* differences in lapse risk
  across segments — which is what a pricing committee typically needs.
- The hazard ratios in Table 3 quantify each covariate's effect:
  values above 1 indicate higher lapse risk relative to the baseline
  category, values below 1 indicate lower risk.
- This completes **Part V — Survival Analysis**. The toolkit —
  censoring (L13), Kaplan-Meier/Nelson-Aalen (L14), parametric models
  (L15), and model comparison (L16) — transfers directly to
  **Part VI — Churn & Retention Modelling**, which begins in Lesson 17.

---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
