---
title: "EDA for Churn & Retention Data"
subtitle: "Advanced R for Actuarial Statistics — Lesson 4 Business Case"
author: "Actuarial Analytics Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: flatly
    number_sections: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 4)
```

---

# Overview

Churn is one of the most commercially significant events in insurance.
Before building any predictive model, actuaries must understand the data
through **Exploratory Data Analysis (EDA)**.

This business case conducts a structured EDA on a 120-customer portfolio,
covering:

- Univariate distributions of key variables
- Churn rate by product type, region, and NPS band
- Tenure analysis — how long before customers leave?
- Bivariate analysis: premium, claims, and NPS vs churn status

> **Data file:** `lesson4_churn_eda.csv` — 120 customer records across Motor,
> Home, and Life products.

---

# Load & Prepare

```{r load}
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)

churn <- read.csv("lesson4_churn_eda.csv",
                  stringsAsFactors = FALSE,
                  na.strings        = c("", "NA"))

churn <- churn %>%
  mutate(
    join_date       = as.Date(join_date),
    churn_date      = as.Date(churn_date),
    last_claim_date = as.Date(last_claim_date),
    obs_date        = as.Date("2024-06-01"),

    # Tenure: months from join to churn (or to obs_date if still active)
    tenure_months   = as.integer(
      interval(join_date,
               if_else(!is.na(churn_date), churn_date, obs_date)) / months(1)
    ),

    churn_label  = if_else(churn_flag == 1, "Churned", "Active"),
    product_type = factor(product_type, levels = c("Motor", "Home", "Life")),

    nps_band = case_when(
      nps_score <= 4 ~ "Detractor (0-4)",
      nps_score <= 7 ~ "Passive (5-7)",
      TRUE           ~ "Promoter (8-10)"
    ),
    nps_band = factor(nps_band,
                      levels = c("Detractor (0-4)", "Passive (5-7)",
                                 "Promoter (8-10)"))
  )

glimpse(churn)
```

---

# Univariate EDA

## Overall Churn Rate

```{r churn-rate}
churn_summary <- churn %>%
  summarise(
    total      = n(),
    n_churned  = sum(churn_flag),
    churn_rate = round(mean(churn_flag) * 100, 1)
  )

knitr::kable(churn_summary,
             col.names = c("Total Customers", "Churned", "Churn Rate (%)"),
             caption   = "Table 1: Overall Churn Rate")
```

```{r fig-status-bar, fig.cap="Figure 1: Active vs Churned Customer Count"}
churn %>%
  count(churn_label) %>%
  ggplot(aes(x = churn_label, y = n, fill = churn_label)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = n), vjust = -0.4, fontface = "bold") +
  scale_fill_manual(values = c("Active" = "#1F4E79", "Churned" = "#C55A11")) +
  labs(title = "Customer Status", x = NULL, y = "Count") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## Premium Distribution

```{r fig-premium, fig.cap="Figure 2: Annual Premium Distribution by Product"}
ggplot(churn, aes(x = premium_annual, fill = product_type)) +
  geom_histogram(bins = 20, colour = "white", alpha = 0.85) +
  facet_wrap(~ product_type, scales = "free_x") +
  scale_fill_manual(values = c("Motor"="#C55A11","Home"="#1F4E79","Life"="#2E86AB")) +
  labs(title = "Annual Premium Distribution by Product Type",
       x = "Annual Premium (£)", y = "Count") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

## Age Distribution by Churn Status

```{r fig-age, fig.cap="Figure 3: Age Distribution — Churned vs Active"}
ggplot(churn, aes(x = age, fill = churn_label)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("Active" = "#1F4E79", "Churned" = "#C55A11")) +
  labs(title = "Age Distribution by Churn Status",
       x = "Customer Age", y = "Density", fill = NULL) +
  theme_minimal(base_size = 12)
```

Churned customers skew younger. This has direct implications for
retention targeting and pricing strategy.

---

# Churn Rate by Segment

## By Product Type

```{r churn-by-product}
product_churn <- churn %>%
  group_by(product_type) %>%
  summarise(
    n_customers = n(),
    n_churned   = sum(churn_flag),
    churn_rate  = round(mean(churn_flag) * 100, 1),
    avg_tenure  = round(mean(tenure_months), 1),
    avg_premium = round(mean(premium_annual), 0),
    .groups = "drop"
  )

knitr::kable(product_churn,
             col.names = c("Product", "Customers", "Churned",
                           "Churn Rate (%)", "Avg Tenure (mo)", "Avg Premium (£)"),
             caption   = "Table 2: Churn Metrics by Product Type")
```

```{r fig-churn-product, fig.cap="Figure 4: Churn Rate by Product Type"}
ggplot(product_churn, aes(x = product_type, y = churn_rate, fill = product_type)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(churn_rate, "%")),
            vjust = -0.4, fontface = "bold", size = 4) +
  scale_fill_manual(values = c("Motor"="#C55A11","Home"="#1F4E79","Life"="#2E86AB")) +
  labs(title = "Churn Rate by Product Type", x = NULL, y = "Churn Rate (%)") +
  ylim(0, max(product_churn$churn_rate) * 1.2) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## By Region

```{r churn-by-region}
region_churn <- churn %>%
  group_by(region) %>%
  summarise(
    n_customers = n(),
    churn_rate  = round(mean(churn_flag) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(churn_rate))

knitr::kable(region_churn,
             col.names = c("Region", "Customers", "Churn Rate (%)"),
             caption   = "Table 3: Churn Rate by Region")
```

## By NPS Band

```{r churn-by-nps}
nps_churn <- churn %>%
  group_by(nps_band) %>%
  summarise(
    n          = n(),
    churn_rate = round(mean(churn_flag) * 100, 1),
    .groups    = "drop"
  )

knitr::kable(nps_churn,
             col.names = c("NPS Band", "Customers", "Churn Rate (%)"),
             caption   = "Table 4: Churn Rate by NPS Band")
```

---

# Tenure Analysis

## Tenure Distribution

```{r fig-tenure, fig.cap="Figure 5: Tenure Distribution by Churn Status"}
ggplot(churn, aes(x = tenure_months, fill = churn_label)) +
  geom_histogram(bins = 15, colour = "white", alpha = 0.85,
                 position = "identity") +
  scale_fill_manual(values = c("Active" = "#1F4E79", "Churned" = "#C55A11")) +
  labs(title = "Tenure Distribution by Churn Status",
       x = "Tenure (months)", y = "Count", fill = NULL) +
  theme_minimal(base_size = 12)
```

## Tenure at Churn by Product

```{r tenure-product}
tenure_summary <- churn %>%
  filter(churn_flag == 1) %>%
  group_by(product_type) %>%
  summarise(
    n             = n(),
    avg_tenure    = round(mean(tenure_months), 1),
    median_tenure = median(tenure_months),
    min_tenure    = min(tenure_months),
    max_tenure    = max(tenure_months),
    .groups = "drop"
  )

knitr::kable(tenure_summary,
             col.names = c("Product", "N Churned", "Mean Tenure (mo)",
                           "Median", "Min", "Max"),
             caption   = "Table 5: Tenure at Churn by Product Type")
```

---

# Bivariate Analysis

## Premium vs Churn

```{r fig-premium-box, fig.cap="Figure 6: Annual Premium by Churn Status"}
ggplot(churn, aes(x = churn_label, y = premium_annual, fill = churn_label)) +
  geom_boxplot(alpha = 0.8, outlier.shape = 21) +
  scale_fill_manual(values = c("Active" = "#1F4E79", "Churned" = "#C55A11")) +
  labs(title = "Annual Premium by Churn Status",
       x = NULL, y = "Annual Premium (£)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## Claims vs Churn Rate

```{r claims-churn}
claims_churn <- churn %>%
  group_by(num_claims, churn_label) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = churn_label, values_from = n,
              values_fill = 0) %>%
  mutate(
    total      = Active + Churned,
    churn_rate = round(Churned / total * 100, 1)
  )

knitr::kable(claims_churn,
             caption = "Table 6: Churn Rate by Number of Claims")
```

## NPS Score vs Tenure

```{r fig-scatter, fig.cap="Figure 7: NPS Score vs Tenure by Churn Status"}
ggplot(churn, aes(x = tenure_months, y = nps_score,
                  colour = churn_label, shape = product_type)) +
  geom_point(alpha = 0.7, size = 2.5) +
  scale_colour_manual(values = c("Active" = "#1F4E79", "Churned" = "#C55A11")) +
  labs(title  = "NPS Score vs Tenure by Churn Status",
       x      = "Tenure (months)", y = "NPS Score",
       colour = "Status", shape = "Product") +
  theme_minimal(base_size = 11)
```

---

# STUDENT EXERCISE

The chunk below is **intentionally incomplete** (`eval=FALSE`).
Change to `eval=TRUE` and fill in the blanks.

**Tasks:**

1. Calculate churn rate by **payment method**
2. Create a bar chart of churn rate by **acquisition channel**
3. Compute average NPS by churn status and use **inline R** in the narrative

```{r student-exercise, eval=FALSE}
# Task 1 — Churn rate by payment method
payment_churn <- churn %>%
  group_by(___) %>%
  summarise(
    n          = n(),
    churn_rate = round(mean(___) * 100, 1)
  )
knitr::kable(payment_churn, caption = "Churn Rate by Payment Method")


# Task 2 — Bar chart by channel
channel_churn <- churn %>%
  group_by(___) %>%
  summarise(churn_rate = round(mean(churn_flag) * 100, 1))

ggplot(channel_churn, aes(x = ___, y = ___, fill = channel)) +
  geom_col(width = 0.5) +
  labs(title = "Churn Rate by Acquisition Channel",
       x = NULL, y = "Churn Rate (%)") +
  theme_minimal()


# Task 3 — Average NPS by churn status
nps_by_status <- churn %>%
  group_by(___) %>%
  summarise(avg_nps = round(mean(___, na.rm = TRUE), 2))

nps_churned <- nps_by_status$avg_nps[nps_by_status$churn_label == "Churned"]
nps_active  <- nps_by_status$avg_nps[nps_by_status$churn_label == "Active"]
```

Churned customers had an average NPS of **`r nps_churned`** vs
**`r nps_active`** for active customers — a gap of
**`r round(nps_active - nps_churned, 2)`** points.

---

# EDA Summary

```{r eda-summary}
overall_churn  <- round(mean(churn$churn_flag) * 100, 1)
motor_churn    <- round(mean(churn$churn_flag[churn$product_type == "Motor"]) * 100, 1)
life_churn     <- round(mean(churn$churn_flag[churn$product_type == "Life"]) * 100, 1)
avg_churn_age  <- round(mean(churn$age[churn$churn_flag == 1]), 1)
avg_active_age <- round(mean(churn$age[churn$churn_flag == 0]), 1)
```

Key findings:

- Overall churn rate: **`r overall_churn`%**
- Motor churn (**`r motor_churn`%**) is substantially higher than Life
  (**`r life_churn`%**), indicating higher price sensitivity
- Churned customers are younger on average (**`r avg_churn_age`** yrs)
  than active customers (**`r avg_active_age`** yrs)
- Lower NPS and credit card payment are associated with higher churn risk
- Most churn occurs within the first 18–30 months of tenure

> **Next step:** Lesson 5 visualises time series and retention curves from this
> same dataset — the patterns you've identified here will become charts.

---

*Generated automatically with R Markdown. Re-knit with updated data to refresh.*
