---
title: "Parametric Survival Models: Exponential & Weibull"
subtitle: "Survival Analysis — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(survival)
library(knitr)

policyholders <- read_csv("L2_policyholders.csv") %>%
  mutate(
    time  = policy_duration_months,
    event = if_else(policy_duration_months < 60, 1, 0)
  )

# KM for comparison
fit_km <- survfit(Surv(time, event) ~ 1, data = policyholders)
```

# Overview

This report fits **Exponential** and **Weibull** parametric survival
models to the policyholder lapse data and compares them against the
non-parametric Kaplan-Meier estimate from Lesson 14.

Portfolio: **`r nrow(policyholders)`** policyholders,
**`r sum(policyholders$event)`** events (lapses),
**`r sum(1 - policyholders$event)`** censored.

---

# Exponential Model

## Fitting

```{r exp-fit, echo=TRUE}
fit_exp <- survreg(Surv(time, event) ~ 1,
                   data  = policyholders,
                   dist  = "exponential")
summary(fit_exp)
```

## Parameter Extraction

```{r exp-params, echo=TRUE}
# survreg returns log(scale); rate lambda = 1 / exp(coef)
lambda_exp <- 1 / exp(coef(fit_exp)[1])

cat("Exponential rate (λ):         ", round(lambda_exp, 5), "\n")
cat("Mean survival time (1/λ):     ",
    round(1 / lambda_exp, 1), "months\n")
cat("Median survival time (-log(0.5)/λ):",
    round(log(2) / lambda_exp, 1), "months\n")
```

## Survival Function

```{r exp-surv}
t_seq    <- seq(0, 72, by = 1)
S_exp    <- exp(-lambda_exp * t_seq)
exp_df   <- tibble(time = t_seq, surv = S_exp, model = "Exponential")
```

---

# Weibull Model

## Fitting

```{r weib-fit, echo=TRUE}
fit_weib <- survreg(Surv(time, event) ~ 1,
                    data = policyholders,
                    dist = "weibull")
summary(fit_weib)
```

## Parameter Extraction

```{r weib-params, echo=TRUE}
# survreg Weibull parameterisation:
# shape k = 1 / fit$scale
# scale λ = exp(coef)
k_weib      <- 1 / fit_weib$scale
lambda_weib <- exp(coef(fit_weib)[1])

cat("Weibull shape (k):  ", round(k_weib, 4), "\n")
cat("Weibull scale (λ):  ", round(lambda_weib, 2), "months\n")
cat("Interpretation:     ",
    ifelse(k_weib < 1,
           "k < 1 → decreasing hazard (early lapses dominate)",
           ifelse(k_weib > 1,
                  "k > 1 → increasing hazard (lapses accelerate over time)",
                  "k = 1 → constant hazard (reduces to Exponential)")),
    "\n")
```

## Survival Function

```{r weib-surv}
S_weib   <- exp(-(t_seq / lambda_weib)^k_weib)
weib_df  <- tibble(time = t_seq, surv = S_weib, model = "Weibull")
```

---

# Model Comparison

## Survival Curves: KM vs Parametric Models

```{r comparison-plot, fig.width=8, fig.height=5,fig.cap="Figure 1: KM vs Exponential vs Weibull Survival Curves"}
km_df <- tibble(
  time  = c(0, fit_km$time),
  surv  = c(1, fit_km$surv),
  model = "Kaplan-Meier"
)

bind_rows(km_df, exp_df, weib_df) %>%
  ggplot(aes(x = time, y = surv, color = model, linetype = model)) +
  geom_step(data = . %>% filter(model == "Kaplan-Meier"),
            linewidth = 0.8) +
  geom_line(data = . %>% filter(model != "Kaplan-Meier"),
            linewidth = 0.8) +
  scale_color_manual(values = c("Kaplan-Meier" = "#2c7bb6",
                                 "Exponential"  = "#d7191c",
                                 "Weibull"      = "#1a9641")) +
  scale_linetype_manual(values = c("Kaplan-Meier" = "solid",
                                    "Exponential"  = "dashed",
                                    "Weibull"      = "dashed")) +
  scale_y_continuous(limits = c(0, 1),
                     labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(0, 72, 12)) +
  labs(title    = "Survival Curves: KM vs Parametric Models",
       subtitle = "Step = Kaplan-Meier | Dashed = parametric fits",
       x = "Months in Force", y = "Survival Probability S(t)",
       color = "Model", linetype = "Model") +
  theme_minimal(base_size = 12)
```

## AIC Comparison

```{r aic-comparison}
tibble(
  Model = c("Exponential", "Weibull"),
  `Log-Likelihood` = round(c(fit_exp$loglik[2],
                               fit_weib$loglik[2]), 2),
  `Parameters`     = c(1, 2),
  AIC              = round(c(AIC(fit_exp), AIC(fit_weib)), 2)
) %>%
  kable(caption = "Table 1: Model Comparison — Log-Likelihood and AIC")
```

```{r aic-conclusion}
best <- if (AIC(fit_weib) < AIC(fit_exp)) "Weibull" else "Exponential"
```

The **`r best`** model has the lower AIC and is preferred.

---

# Hazard Functions

```{r hazard-plot, fig.width=8, fig.height=4,fig.cap="Figure 2: Fitted Hazard Functions — Exponential vs Weibull"}
h_exp  <- rep(lambda_exp, length(t_seq))
h_weib <- (k_weib / lambda_weib) * (t_seq / lambda_weib)^(k_weib - 1)

tibble(
  time  = rep(t_seq, 2),
  hazard = c(h_exp, h_weib),
  model  = rep(c("Exponential", "Weibull"), each = length(t_seq))
) %>%
  filter(time > 0) %>%
  ggplot(aes(x = time, y = hazard, color = model)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("Exponential" = "#d7191c",
                                 "Weibull"     = "#1a9641")) +
  labs(title = "Hazard Function h(t) — Exponential vs Weibull",
       x = "Months in Force", y = "Hazard Rate h(t)",
       color = "Model") +
  theme_minimal(base_size = 12)
```

---

# Survival Estimates at Key Durations

```{r survival-table}
times_key <- c(12, 24, 36, 48, 60)

tibble(
  `Time (Months)` = times_key,
  `Kaplan-Meier`  = paste0(
    round(summary(fit_km, times = times_key)$surv * 100, 1), "%"),
  `Exponential`   = paste0(
    round(exp(-lambda_exp * times_key) * 100, 1), "%"),
  `Weibull`       = paste0(
    round(exp(-(times_key / lambda_weib)^k_weib) * 100, 1), "%")
) %>%
  kable(caption = "Table 2: Survival Estimates at Key Durations")
```

---

# Key Findings

- The **Exponential model** assumes a constant hazard rate of
  **λ = `r round(lambda_exp, 5)`** per month — meaning the lapse
  rate does not change with policy duration. Mean survival time:
  **`r round(1/lambda_exp, 1)` months**.
- The **Weibull shape parameter k = `r round(k_weib, 3)`** indicates
  `r ifelse(k_weib < 1, "a decreasing hazard — lapses are most likely early in the policy life and become less frequent over time.", ifelse(k_weib > 1, "an increasing hazard — lapse risk grows as the policy matures.", "a constant hazard — equivalent to the Exponential model."))`
- The **`r best`** model is preferred by AIC
  (ΔAIC = `r round(abs(AIC(fit_exp) - AIC(fit_weib)), 2)`).


---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
