---
title: "Visualization for Time Series & Retention Curves"
subtitle: "Advanced R for Actuarial Statistics — Lesson 5 Business Case"
author: "Actuarial Analytics Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: flatly
    number_sections: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

---

# Overview

This business case builds on the EDA from Lesson 4. Having identified *what*
the churn patterns are, we now focus on *when* and *how* they evolve over time.

| File | Content |
|---|---|
| `lesson5_monthly_ts.csv` | 36-month aggregated portfolio metrics (Jan 2022–Dec 2024) |
| `lesson5_cohort_retention.csv` | 8 quarterly cohorts tracked over 12 months |
| `lesson4_churn_eda.csv` | Individual-level data for cohort construction demo |

> Place all three CSV files in your working directory before knitting.

---

# Setup

```{r load}
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(scales)

ts_data <- read.csv("lesson5_monthly_ts.csv", stringsAsFactors = FALSE) %>%
  mutate(
    month        = as.Date(paste0(month, "-01")),
    product_type = factor(product_type, levels = c("Motor", "Home", "Life"))
  )

cohort <- read.csv("lesson5_cohort_retention.csv", stringsAsFactors = FALSE) %>%
  mutate(
    cohort_quarter = factor(cohort_quarter,
                            levels = c("Q1-2022","Q2-2022","Q3-2022","Q4-2022",
                                       "Q1-2023","Q2-2023","Q3-2023","Q4-2023"))
  )

churn <- read.csv("lesson4_churn_eda.csv", stringsAsFactors = FALSE,
                  na.strings = c("", "NA")) %>%
  mutate(
    join_date    = as.Date(join_date),
    churn_date   = as.Date(churn_date),
    obs_date     = as.Date("2024-06-01"),
    churn_label  = if_else(churn_flag == 1, "Churned", "Active"),
    product_type = factor(product_type, levels = c("Motor", "Home", "Life"))
  )

glimpse(ts_data)
```

---

# Time Series: Portfolio Growth

## Multi-Line Chart

```{r fig-active-line, fig.cap="Figure 1: Active Policy Count Over Time by Product"}
ggplot(ts_data, aes(x = month, y = active_policies,
                    colour = product_type, group = product_type)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8, alpha = 0.7) +
  scale_colour_manual(values = c("Motor"="#C55A11","Home"="#1F4E79","Life"="#2E86AB")) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  scale_y_continuous(labels = comma) +
  labs(title  = "Active Policy Count — Jan 2022 to Dec 2024",
       x = NULL, y = "Active Policies", colour = "Product") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        plot.title   = element_text(face = "bold"))
```

## Annotated Chart

```{r fig-annotated, fig.cap="Figure 2: Motor Policy Count with Event Annotation"}
motor_ts         <- ts_data %>% filter(product_type == "Motor")
rate_change_date <- as.Date("2023-01-01")

ggplot(motor_ts, aes(x = month, y = active_policies)) +
  geom_ribbon(aes(ymin = active_policies * 0.97,
                  ymax = active_policies * 1.03),
              fill = "#C55A11", alpha = 0.15) +
  geom_line(colour = "#C55A11", linewidth = 1.2) +
  geom_vline(xintercept = as.numeric(rate_change_date),
             linetype = "dashed", colour = "grey40", linewidth = 0.8) +
  annotate("text", x = rate_change_date + 20,
           y = max(motor_ts$active_policies) * 0.92,
           label = "Rate\nChange", hjust = 0, size = 3.5, colour = "grey30") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  scale_y_continuous(labels = comma) +
  labs(title = "Motor: Active Policies with Confidence Band",
       x = NULL, y = "Active Policies") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        plot.title   = element_text(face = "bold"))
```

---

# Seasonality: Claim Rate Heatmap

```{r fig-heatmap, fig.cap="Figure 3: Claim Rate Heatmap by Month and Product"}
ts_data %>%
  mutate(yr = year(month), mon = month(month, label = TRUE, abbr = TRUE)) %>%
  ggplot(aes(x = mon, y = factor(yr), fill = claim_rate)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  facet_wrap(~ product_type) +
  scale_fill_gradient2(low = "#BDD7EE", mid = "#1F4E79", high = "#C55A11",
                       midpoint = 0.08,
                       labels = percent_format(accuracy = 0.1)) +
  labs(title = "Claim Rate Heatmap — by Month, Year, and Product",
       x = NULL, y = NULL, fill = "Claim Rate") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        strip.text  = element_text(face = "bold"))
```

Motor claims peak in November–February — consistent with winter driving
conditions. Life shows almost no seasonality, as expected actuarially.

## LOESS Trend on Churn Rate

```{r fig-churn-trend, fig.cap="Figure 4: Monthly Churn Rate with Smoothed Trend"}
ts_data %>%
  mutate(churn_rate_pct = churned_policies / active_policies * 100) %>%
  ggplot(aes(x = month, y = churn_rate_pct,
             colour = product_type, group = product_type)) +
  geom_line(alpha = 0.4, linewidth = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.3, span = 0.4) +
  scale_colour_manual(values = c("Motor"="#C55A11","Home"="#1F4E79","Life"="#2E86AB")) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(title    = "Monthly Churn Rate with LOESS Smoothing",
       subtitle  = "Thin = raw rate; thick = smoothed trend",
       x = NULL, y = "Churn Rate (%)", colour = "Product") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        plot.title   = element_text(face = "bold"))
```

---

# Cohort Retention Curves

## All Cohorts

```{r fig-cohort-all, fig.cap="Figure 5: Retention Curves — All 8 Cohorts"}
ggplot(cohort, aes(x = month, y = retention_rate,
                   colour = cohort_quarter, group = cohort_quarter)) +
  geom_line(linewidth = 1.0, alpha = 0.85) +
  geom_point(size = 1.5) +
  scale_colour_manual(values = c(
    "Q1-2022"="#1F4E79","Q2-2022"="#2E75B6","Q3-2022"="#2E86AB","Q4-2022"="#70AD47",
    "Q1-2023"="#FFC000","Q2-2023"="#ED7D31","Q3-2023"="#C55A11","Q4-2023"="#7030A0"
  )) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0.55, 1.01)) +
  scale_x_continuous(breaks = 1:12) +
  labs(title    = "Retention Curves by Quarterly Cohort",
       subtitle  = "Month 1 = 100% by definition",
       x = "Months Since Joining", y = "Retention Rate", colour = "Cohort") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## 2022 vs 2023 Cohort Comparison

```{r fig-cohort-compare, fig.cap="Figure 6: Average Retention — 2022 vs 2023 Cohorts"}
cohort %>%
  mutate(era = if_else(cohort_quarter %in% c("Q1-2022","Q2-2022","Q3-2022","Q4-2022"),
                       "2022 Cohorts", "2023 Cohorts")) %>%
  group_by(era, month) %>%
  summarise(avg_ret = mean(retention_rate), .groups = "drop") %>%
  ggplot(aes(x = month, y = avg_ret, colour = era, fill = era, group = era)) +
  geom_ribbon(aes(ymin = avg_ret - 0.02, ymax = avg_ret + 0.02),
              alpha = 0.12, colour = NA) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c("2022 Cohorts"="#1F4E79","2023 Cohorts"="#C55A11")) +
  scale_fill_manual(values   = c("2022 Cohorts"="#1F4E79","2023 Cohorts"="#C55A11")) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0.55, 1.01)) +
  scale_x_continuous(breaks = 1:12) +
  labs(title    = "Average Retention: 2022 vs 2023 Cohorts",
       subtitle  = "Shaded band = ±2 pp",
       x = "Months Since Joining", y = "Average Retention Rate",
       colour = NULL, fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Cohort Retention Heatmap

```{r fig-cohort-heatmap, fig.cap="Figure 7: Cohort Retention Heatmap"}
ggplot(cohort, aes(x = factor(month), y = cohort_quarter, fill = retention_rate)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = scales::percent(retention_rate, accuracy = 1)),
            size = 2.8, colour = "white", fontface = "bold") +
  scale_fill_gradient2(low = "#C55A11", mid = "#1F4E79", high = "#BDD7EE",
                       midpoint = 0.78,
                       labels = percent_format(accuracy = 1)) +
  scale_x_discrete(labels = paste0("M", 1:12)) +
  labs(title = "Cohort Retention Heatmap",
       x = "Month Since Joining", y = "Cohort", fill = "Retention") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.y = element_text(size = 9))
```

---

# Build Cohort Table from Individual Data

```{r build-cohort}
churn_cohort <- churn %>%
  mutate(
    cohort = case_when(
      join_date < as.Date("2019-07-01") ~ "H1-2019",
      join_date < as.Date("2020-01-01") ~ "H2-2019",
      join_date < as.Date("2020-07-01") ~ "H1-2020",
      join_date < as.Date("2021-01-01") ~ "H2-2020",
      join_date < as.Date("2021-07-01") ~ "H1-2021",
      join_date < as.Date("2022-01-01") ~ "H2-2021",
      TRUE                              ~ "H1-2022+"
    )
  )

retention_milestones <- churn_cohort %>%
  mutate(
    tenure_months = as.integer(
      interval(join_date, if_else(!is.na(churn_date), churn_date, obs_date))
      / months(1)
    )
  ) %>%
  group_by(cohort) %>%
  summarise(
    size     = n(),
    ret_12mo = round(mean(tenure_months >= 12 | churn_flag == 0) * 100, 1),
    ret_24mo = round(mean(tenure_months >= 24 | churn_flag == 0) * 100, 1),
    ret_36mo = round(mean(tenure_months >= 36 | churn_flag == 0) * 100, 1),
    .groups  = "drop"
  )

knitr::kable(retention_milestones,
             col.names = c("Cohort", "Size", "Ret @ 12mo (%)",
                           "Ret @ 24mo (%)", "Ret @ 36mo (%)"),
             caption   = "Table 1: Retention at Key Milestones by Cohort")
```

---

# STUDENT EXERCISE

Change `eval=FALSE` to `eval=TRUE` and complete the blanks.

**Tasks:**

1. Faceted line chart of `total_premium_gbp` over time — one panel per `product_type`
2. Smoothed trend line on Motor active policy count using `geom_smooth()`
3. Horizontal bar chart of 12-month retention rate, one bar per cohort, ordered highest to lowest

```{r student-exercise, eval=FALSE}
# Task 1 — Faceted premium time series
ggplot(ts_data, aes(x = ___, y = total_premium_gbp, colour = product_type)) +
  geom_line(linewidth = 1.1) +
  facet_wrap(~ ___, scales = "free_y") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = ___) +
  labs(title = "Total Premium GWP Over Time", x = NULL, y = "Total Premium (£)") +
  theme_minimal() + theme(legend.position = "none")


# Task 2 — Motor with smoothed trend
motor_only <- ts_data %>% filter(___ == "Motor")

ggplot(motor_only, aes(x = month, y = ___)) +
  geom_line(colour = "#C55A11", alpha = 0.5) +
  geom_smooth(method = "___", se = TRUE, colour = "#C55A11") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(title = "Motor: Active Policy Count with Trend",
       x = NULL, y = "Active Policies") +
  theme_minimal()


# Task 3 — Month-12 retention bar chart
ret_12 <- cohort %>%
  filter(month == ___) %>%
  select(cohort_quarter, retention_rate)

ggplot(ret_12, aes(x = reorder(cohort_quarter, retention_rate),
                   y = ___,
                   fill = retention_rate)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_gradient(low = "#C55A11", high = "#1F4E79") +
  labs(title = "12-Month Retention Rate by Cohort",
       x = NULL, y = "Retention Rate") +
  theme_minimal() + theme(legend.position = "none")
```

---

# Key Findings

```{r key-findings}
avg_12mo_ret <- cohort %>%
  filter(month == 12) %>%
  summarise(avg = round(mean(retention_rate) * 100, 1)) %>%
  pull(avg)

motor_peak_churn <- ts_data %>%
  filter(product_type == "Motor") %>%
  mutate(cr = churned_policies / active_policies) %>%
  arrange(desc(cr)) %>%
  slice(1) %>%
  mutate(label = format(month, "%B %Y")) %>%
  pull(label)

total_gwp_dec24 <- ts_data %>%
  filter(month == as.Date("2024-12-01")) %>%
  summarise(total = sum(total_premium_gbp)) %>%
  pull(total)
```

- Average **12-month retention across all cohorts: `r avg_12mo_ret`%** —
  roughly 1 in 4 new customers does not reach first renewal.
- Highest Motor churn month: **`r motor_peak_churn`** — consistent with
  winter renewal pressure and price comparison activity.
- Total portfolio GWP at December 2024:
  **£`r format(total_gwp_dec24, big.mark=",")`**.
- 2023 cohorts show systematically lower 12-month retention than 2022 —
  a signal of deteriorating new business quality or channel mix shift.

> **Next step:** Lessons 6–9 introduce formal time series modelling —
> `ts`, `xts`, ACF/PACF, and ARIMA — adding statistical rigour to
> the patterns visualised here.

---

*Generated automatically with R Markdown. Re-knit with updated data to refresh.*
