---
title: "Kaplan-Meier & Nelson-Aalen Estimators"
subtitle: "Survival Analysis — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(survival)
library(survminer)
library(knitr)

policyholders <- read_csv("L2_policyholders.csv") %>%
  mutate(
    time  = policy_duration_months,
    event = if_else(policy_duration_months < 60, 1, 0)
  )
```

# Overview

This report estimates and compares survival curves across policy types
using the **Kaplan-Meier** and **Nelson-Aalen** estimators, and formally
tests whether observed differences are statistically significant using
the **log-rank test**.

Portfolio: **`r nrow(policyholders)`** policyholders,
**`r sum(policyholders$event)`** events (lapses),
**`r sum(1 - policyholders$event)`** censored.

---

# Kaplan-Meier Estimator

## Overall KM Curve

```{r km-fit, echo=TRUE}
fit_km <- survfit(Surv(time, event) ~ 1, data = policyholders)
```

```{r km-summary-table}
km_summary <- summary(fit_km, times = c(6, 12, 24, 36, 48, 60))

tibble(
  `Time (Months)`    = km_summary$time,
  `At Risk`          = km_summary$n.risk,
  `Events`           = km_summary$n.event,
  `S(t)`             = round(km_summary$surv,  3),
  `Lower 95% CI`     = round(km_summary$lower, 3),
  `Upper 95% CI`     = round(km_summary$upper, 3)
) %>%
  kable(caption = "Table 1: Kaplan-Meier Survival Estimates at Key Durations")
```

```{r km-overall-plot, fig.width=8, fig.height=5,fig.cap="Figure 1: Kaplan-Meier Curve — Overall Portfolio"}
ggsurvplot(
  fit_km,
  data          = policyholders,
  conf.int      = TRUE,
  risk.table    = TRUE,
  palette       = "#2c7bb6",
  xlab          = "Months in Force",
  ylab          = "Survival Probability S(t)",
  title         = "Kaplan-Meier — Overall Portfolio",
  ggtheme       = theme_minimal(base_size = 12),
  risk.table.height = 0.25
)
```

## KM by Policy Type

```{r km-by-type, echo=TRUE}
fit_type <- survfit(Surv(time, event) ~ policy_type,
                    data = policyholders)
```

```{r km-type-plot, fig.width=8, fig.height=6,fig.cap="Figure 2: Kaplan-Meier Curves by Policy Type"}
ggsurvplot(
  fit_type,
  data          = policyholders,
  conf.int      = TRUE,
  risk.table    = TRUE,
  pval          = TRUE,
  palette       = "Set1",
  legend.title  = "Policy Type",
  xlab          = "Months in Force",
  ylab          = "Survival Probability S(t)",
  title         = "Kaplan-Meier by Policy Type",
  ggtheme       = theme_minimal(base_size = 11),
  risk.table.height = 0.28
)
```

---

# Nelson-Aalen Estimator

```{r na-fit, echo=TRUE}
fit_na <- survfit(Surv(time, event) ~ 1,
                  data = policyholders,
                  type = "fh")   # Fleming-Harrington = Nelson-Aalen

na_df <- tibble(
  time  = fit_na$time,
  cumhaz = -log(fit_na$surv)    # H(t) = -log(S(t))
)
```

```{r na-km-compare-plot, fig.width=8,fig.height=4,fig.cap="Figure 3: Cumulative Hazard — Nelson-Aalen vs KM-derived"}
km_df <- tibble(
  time   = fit_km$time,
  cumhaz = -log(fit_km$surv)
)

bind_rows(
  na_df %>% mutate(method = "Nelson-Aalen"),
  km_df %>% mutate(method = "KM-derived H(t) = -log(S(t))")
) %>%
  ggplot(aes(x = time, y = cumhaz, color = method)) +
  geom_step(linewidth = 0.8) +
  scale_color_manual(values = c("Nelson-Aalen"           = "#d7191c",
                                 "KM-derived H(t) = -log(S(t))" = "#2c7bb6")) +
  labs(title  = "Cumulative Hazard Function H(t)",
       x = "Months in Force", y = "Cumulative Hazard H(t)",
       color = NULL) +
  theme_minimal(base_size = 12)
```

---

# Log-Rank Test

```{r logrank, echo=TRUE}
lr_test <- survdiff(Surv(time, event) ~ policy_type,
                    data = policyholders)
lr_test
```

```{r logrank-result}
p_val <- 1 - pchisq(lr_test$chisq, df = length(lr_test$n) - 1)

tibble(
  `Chi-squared statistic` = round(lr_test$chisq, 3),
  `Degrees of freedom`    = length(lr_test$n) - 1,
  `p-value`               = round(p_val, 4),
  `Conclusion`            = ifelse(
    p_val < 0.05,
    "Significant difference across policy types (p < 0.05)",
    "No significant difference across policy types (p ≥ 0.05)"
  )
) %>%
  kable(caption = "Table 2: Log-Rank Test Results")
```

---

# Observed vs Expected Events

```{r obs-exp}
tibble(
  `Policy Type` = names(lr_test$n),
  `N`           = lr_test$n,
  `Observed`    = lr_test$obs,
  `Expected`    = round(lr_test$exp, 1),
  `(O-E)²/E`   = round((lr_test$obs - lr_test$exp)^2 / lr_test$exp, 3)
) %>%
  kable(caption = "Table 3: Observed vs Expected Events by Policy Type")
```

---

# Key Findings

```{r kf-vals, include=FALSE}
p_val_fmt <- round(p_val, 4)
km_12 <- round(summary(fit_km, times = 12)$surv * 100, 1)
km_24 <- round(summary(fit_km, times = 24)$surv * 100, 1)
```

- The overall **12-month survival rate** is **`r km_12`%** and the
  **24-month survival rate** is **`r km_24`%** — these are the
  portfolio-level retention rates at key pricing durations.
- The **log-rank test** yields p =
  **`r p_val_fmt`** —
  `r ifelse(p_val < 0.05, "survival curves differ significantly across policy types, justifying product-level lapse assumptions in pricing models.", "no statistically significant difference across policy types is detected at the 5% level.")`
- The **Nelson-Aalen** and KM-derived cumulative hazard estimates track
  closely — as expected for large samples. Nelson-Aalen is preferred
  when estimating the hazard function itself; KM is preferred for
  the survival probability S(t).

---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
