---
title: "Poisson & Compound Poisson Simulation"
subtitle: "Stochastic Modelling — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(knitr)

set.seed(42)

policyholders <- read_csv("L2_policyholders.csv")

# Estimate lambda: average monthly claim count from portfolio
# Portfolio spans ~4 years (48 months) across 1000 policyholders
total_claims <- sum(policyholders$has_claim)
n_months     <- 48
lambda       <- total_claims / n_months

# Estimate severity: lognormal parameters from observed claim amounts
claims_only  <- policyholders$claim_amount[policyholders$has_claim == 1]
mu_sev       <- mean(log(claims_only))
sigma_sev    <- sd(log(claims_only))
```

# Overview

This report simulates claim arrivals and aggregate losses using **Poisson**
and **Compound Poisson** processes, calibrated from the
**`r nrow(policyholders)`-policyholder** portfolio.

Estimated parameters from portfolio data:

- **Claim frequency** (λ): **`r round(lambda, 2)` claims per month**
  (`r total_claims` total claims over `r n_months` months)
- **Claim severity**: lognormal with
  μ = `r round(mu_sev, 3)`, σ = `r round(sigma_sev, 3)`
  (mean severity = **£`r format(round(mean(claims_only), 0), big.mark=",")`**)

---

# Poisson Process: Claim Arrivals

## Single Month Simulation

```{r single-month, echo=TRUE}
lambda   <- total_claims / n_months
n_claims <- rpois(1, lambda = lambda)
cat("Simulated claims this month:", n_claims, "\n")
```

## Monthly Claim Counts Over 36 Months

```{r monthly-counts, echo=TRUE}
n_months_sim <- 36
monthly_claims <- rpois(n_months_sim, lambda = lambda)
cumulative_claims <- cumsum(monthly_claims)
```

```{r poisson-plot, fig.width=8, fig.height=4,fig.cap="Figure 1: Simulated Monthly and Cumulative Claim Counts"}
tibble(
  month      = 1:n_months_sim,
  monthly    = monthly_claims,
  cumulative = cumulative_claims
) %>%
  pivot_longer(-month, names_to = "type", values_to = "count") %>%
  mutate(type = recode(type,
                       monthly    = "Monthly Count",
                       cumulative = "Cumulative Count")) %>%
  ggplot(aes(x = month, y = count, color = type)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5, alpha = 0.6) +
  scale_color_manual(values = c("Monthly Count"    = "#2c7bb6",
                                 "Cumulative Count" = "#d7191c")) +
  labs(title  = "Simulated Claim Counts Over 36 Months",
       x      = "Month", y = "Claim Count", color = NULL) +
  theme_minimal(base_size = 12)
```

## Poisson Distribution Check

```{r poisson-dist, fig.width=7, fig.height=3.5,fig.cap="Figure 2: Distribution of Monthly Claim Counts vs Theoretical Poisson"}
x_vals <- min(monthly_claims):max(monthly_claims)

theoretical <- tibble(
  x     = x_vals,
  count = dpois(x_vals, lambda) * n_months_sim
)

tibble(count = monthly_claims) %>%
  ggplot(aes(x = count)) +
  geom_bar(fill = "#2c7bb6", alpha = 0.8) +
  geom_point(data = theoretical,
             aes(x = x, y = count),
             color = "#d7191c", size = 3) +
  labs(title    = "Monthly Claim Counts vs Theoretical Poisson(λ)",
       subtitle = "Bars = simulated, Red points = theoretical",
       x = "Claims per Month", y = "Frequency") +
  theme_minimal(base_size = 12)
```

---

# Compound Poisson: Aggregate Losses

## Single Month Aggregate Loss

```{r compound-single, echo=TRUE}
# Step 1: draw number of claims
n  <- rpois(1, lambda = lambda)

# Step 2: draw severities for each claim
severities <- rlnorm(n, meanlog = mu_sev, sdlog = sigma_sev)

# Step 3: aggregate
S <- sum(severities)
cat("Claims this month:", n, "\n")
cat("Aggregate loss:   £", format(round(S, 0), big.mark = ","), "\n")
```

## 36-Month Aggregate Loss Simulation

```{r compound-monthly, echo=TRUE}
aggregate_losses <- sapply(1:n_months_sim, function(t) {
  n <- rpois(1, lambda = lambda)
  sum(rlnorm(n, meanlog = mu_sev, sdlog = sigma_sev))
})

cumulative_losses <- cumsum(aggregate_losses)
```

```{r compound-plot, fig.width=8, fig.height=4,fig.cap="Figure 3: Simulated Monthly and Cumulative Aggregate Losses"}
tibble(
  month      = 1:n_months_sim,
  monthly    = aggregate_losses,
  cumulative = cumulative_losses
) %>%
  pivot_longer(-month, names_to = "type", values_to = "loss") %>%
  mutate(type = recode(type,
                       monthly    = "Monthly Loss",
                       cumulative = "Cumulative Loss")) %>%
  ggplot(aes(x = month, y = loss, color = type)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("Monthly Loss"    = "#2c7bb6",
                                 "Cumulative Loss" = "#d7191c")) +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  labs(title = "Simulated Aggregate Losses Over 36 Months",
       x = "Month", y = "Loss Amount (£)", color = NULL) +
  theme_minimal(base_size = 12)
```

---

# Monte Carlo: Distribution of Annual Aggregate Loss

```{r monte-carlo, echo=TRUE}
n_sims       <- 5000
annual_losses <- replicate(n_sims, {
  monthly <- sapply(1:12, function(t) {
    n <- rpois(1, lambda = lambda)
    sum(rlnorm(n, meanlog = mu_sev, sdlog = sigma_sev))
  })
  sum(monthly)
})
```

```{r mc-dist-plot, fig.width=8, fig.height=4,fig.cap="Figure 4: Distribution of Annual Aggregate Loss (5,000 Simulations)"}
q95 <- quantile(annual_losses, 0.95)
q99 <- quantile(annual_losses, 0.99)

tibble(loss = annual_losses) %>%
  ggplot(aes(x = loss)) +
  geom_histogram(bins = 60, fill = "#2c7bb6",
                 color = "white", alpha = 0.8) +
  geom_vline(xintercept = q95, color = "#fdae61",
             linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = q99, color = "#d7191c",
             linetype = "dashed", linewidth = 0.8) +
  scale_x_continuous(labels = scales::comma_format(prefix = "£")) +
  labs(title    = "Distribution of Annual Aggregate Loss",
       subtitle = "Orange = 95th percentile | Red = 99th percentile",
       x = "Annual Aggregate Loss (£)", y = "Count") +
  theme_minimal(base_size = 12)
```

```{r mc-table}
tibble(
  Statistic       = c("Mean", "Std Dev", "50th (Median)",
                       "95th Percentile", "99th Percentile",
                       "99.5th Percentile"),
  `Value (£)`     = format(
    round(c(mean(annual_losses),
            sd(annual_losses),
            quantile(annual_losses, 0.50),
            quantile(annual_losses, 0.95),
            quantile(annual_losses, 0.99),
            quantile(annual_losses, 0.995)), 0),
    big.mark = ",")
) %>%
  kable(caption = "Table 1: Annual Aggregate Loss — Summary Statistics")
```

---

# Key Findings

```{r kf-vals, include=FALSE}
mean_annual  <- round(mean(annual_losses), 0)
p95_annual   <- round(quantile(annual_losses, 0.95), 0)
p995_annual  <- round(quantile(annual_losses, 0.995), 0)
```

- The expected annual aggregate loss is
  **£`r format(mean_annual, big.mark=",")`**, consistent with
  λ × E[severity] × 12 months.
- The **95th percentile** of annual losses is
  **£`r format(p95_annual, big.mark=",")`** — the typical reserve
  buffer target under a 1-in-20 stress.
- The **99.5th percentile** is
  **£`r format(p995_annual, big.mark=",")`** — the Solvency Capital Requirement.


---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
