---
title: "Intro to Stochastic Processes"
subtitle: "Stochastic Modelling — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(knitr)

set.seed(42)

policyholders <- read_csv("L2_policyholders.csv")

# Parameters from data
n_steps  <- 60          # months of simulation
n_paths  <- 200         # number of simulated paths
mu       <- mean(policyholders$claim_amount[policyholders$has_claim == 1])
sigma    <- sd(policyholders$claim_amount[policyholders$has_claim == 1])
```

# Overview

This report introduces stochastic process simulation in the context of
actuarial claim modelling. Using parameters estimated from the
**`r nrow(policyholders)`-policyholder** portfolio, we simulate
**`r n_paths` independent random walk paths** over **`r n_steps` months**
to illustrate the probabilistic range of cumulative claim outcomes.

Estimated monthly claim increment parameters from the portfolio:
**mean = £`r format(round(mu, 0), big.mark=",")`**,
**sd = £`r format(round(sigma, 0), big.mark=",")`**.

---

# Single Random Walk

```{r single-rw, echo=TRUE}
n_steps <- 60

increments <- rnorm(n_steps, mean = mu, sd = sigma)
rw_path    <- cumsum(increments)
```

```{r single-rw-plot, fig.width=8, fig.height=4,fig.cap="Figure 1: Single Random Walk — Cumulative Claim Amount"}
tibble(
  step   = 1:n_steps,
  value  = rw_path
) %>%
  ggplot(aes(x = step, y = value)) +
  geom_line(color = "#2c7bb6", linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  labs(title = "Single Random Walk — Cumulative Claim Amount",
       x = "Month", y = "Cumulative Amount (£)") +
  theme_minimal(base_size = 12)
```

---

# Multiple Paths

```{r multi-paths, echo=TRUE}
n_paths <- 200

sim_matrix <- replicate(
  n_paths,
  cumsum(rnorm(n_steps, mean = mu, sd = sigma))
)
```

```{r multi-paths-plot, fig.width=8, fig.height=5,fig.cap="Figure 2: 200 Simulated Random Walk Paths"}
sim_df <- as.data.frame(sim_matrix) %>%
  mutate(step = 1:n_steps) %>%
  pivot_longer(-step, names_to = "path", values_to = "value")

ggplot(sim_df, aes(x = step, y = value, group = path)) +
  geom_line(alpha = 0.07, color = "#2c7bb6") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  labs(title = "200 Simulated Paths — Cumulative Claim Amount",
       x = "Month", y = "Cumulative Amount (£)") +
  theme_minimal(base_size = 12)
```

---

# Uncertainty Quantification

```{r uncertainty, echo=TRUE}
final_values <- sim_matrix[n_steps, ]

quantiles <- quantile(final_values,
                      probs = c(0.05, 0.25, 0.50, 0.75, 0.95))
```

```{r quantile-table}
tibble(
  Percentile      = c("5th", "25th", "50th (Median)", "75th", "95th"),
  `Value (£)`     = format(round(quantiles, 0), big.mark = ",")
) %>%
  kable(caption = "Table 1: Distribution of Cumulative Claims at Month 60")
```

```{r distribution-plot, fig.width=8, fig.height=4,fig.cap="Figure 3: Distribution of Cumulative Claim Amount at Month 60"}
tibble(final = final_values) %>%
  ggplot(aes(x = final)) +
  geom_histogram(bins = 40, fill = "#2c7bb6",
                 color = "white", alpha = 0.8) +
  geom_vline(xintercept = quantiles["50%"],
             color = "#d7191c", linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = quantiles["5%"],
             color = "#fdae61", linetype = "dashed", linewidth = 0.7) +
  geom_vline(xintercept = quantiles["95%"],
             color = "#fdae61", linetype = "dashed", linewidth = 0.7) +
  scale_x_continuous(labels = scales::comma_format(prefix = "£")) +
  labs(title = "Distribution of Cumulative Claims at Month 60",
       subtitle = "Red = median, Orange = 5th/95th percentiles",
       x = "Cumulative Claim Amount (£)", y = "Count") +
  theme_minimal(base_size = 12)
```

---

# Paths with Confidence Band

```{r confidence-band, fig.width=8, fig.height=5,fig.cap="Figure 4: Simulated Paths with 90% Confidence Band"}
band_df <- sim_df %>%
  group_by(step) %>%
  summarise(
    median = median(value),
    lo90   = quantile(value, 0.05),
    hi90   = quantile(value, 0.95),
    .groups = "drop"
  )

ggplot() +
  geom_line(data = sim_df,
            aes(x = step, y = value, group = path),
            alpha = 0.05, color = "#2c7bb6") +
  geom_ribbon(data = band_df,
              aes(x = step, ymin = lo90, ymax = hi90),
              fill = "#fdae61", alpha = 0.4) +
  geom_line(data = band_df,
            aes(x = step, y = median),
            color = "#d7191c", linewidth = 0.9) +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  labs(title = "Simulated Paths with 90% Confidence Band",
       subtitle = "Red = median path, Orange band = 5th–95th percentile range",
       x = "Month", y = "Cumulative Amount (£)") +
  theme_minimal(base_size = 12)
```

---

# Key Findings

- A single random walk path is unpredictable — each simulation produces a
  different outcome. This is the fundamental nature of stochastic processes.
- Across **`r n_paths` simulated paths**, the 90% confidence band at month
  **`r n_steps`** runs from
  **£`r format(round(quantiles["5%"], 0), big.mark=",")`** to
  **£`r format(round(quantiles["95%"], 0), big.mark=",")`**.
- The median cumulative claim at month 60 is
  **£`r format(round(quantiles["50%"], 0), big.mark=",")`**.
- This range quantifies the **reserving uncertainty** an actuary must account
  for when setting aside capital for future claims — a direct IFoA CS2
  application.
- In **Lesson 11**, we move from random walks to Markov chains, where the
  next state depends on the current state in a structured way.

---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
