---
title: "Feature Engineering for Churn"
subtitle: "Applied Predictive & Causal Analytics — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(knitr)

policyholders <- read_csv("L2_policyholders.csv") %>%
  mutate(
    policy_start_date = as.Date(policy_start_date),
    churned = if_else(policy_duration_months < 24, 1, 0)
  )
```

# Overview

This report engineers a model-ready feature set for churn prediction from
the **`r nrow(policyholders)`-policyholder** portfolio. The output feeds
directly into the GLM (Lesson 22) and machine-learning models (Lesson 23).

Target variable: **churned** (1 if tenure < 24 months).
Base churn rate: **`r paste0(round(mean(policyholders$churned)*100,1),"%")`**.

---

# Step 1 — Derived Numeric Features

```{r derived-numeric, echo=TRUE}
features <- policyholders %>%
  mutate(
    # Ratio feature: claim amount per premium pound
    claim_to_premium = claim_amount / annual_premium,
    # Premium per month of tenure
    premium_per_month = annual_premium / pmax(policy_duration_months, 1),
    # Binary: high-value customer flag
    high_value = if_else(annual_premium > median(annual_premium), 1, 0)
  )
```

```{r derived-table}
features %>%
  select(policyholder_id, annual_premium, claim_amount,
         claim_to_premium, premium_per_month, high_value) %>%
  head(6) %>%
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kable(caption = "Table 1: Derived Numeric Features (first 6 rows)")
```

---

# Step 2 — Categorical Encoding

```{r categorical, echo=TRUE}
features <- features %>%
  mutate(
    age_band = cut(age,
                   breaks = c(0, 35, 50, 65, 100),
                   labels = c("Under35", "Age35_50", "Age50_65", "Over65")),
    # Ensure factor type for modelling
    policy_type = factor(policy_type),
    region      = factor(region),
    gender      = factor(gender)
  )
```

```{r dummy-demo, echo=TRUE}
# One-hot (dummy) encoding preview using model.matrix
dummy_preview <- model.matrix(~ policy_type - 1,
                              data = features) %>%
  as_tibble() %>%
  head(5)
```

```{r dummy-table}
dummy_preview %>%
  kable(caption = "Table 2: One-Hot Encoding of policy_type (first 5 rows)")
```

---

# Step 3 — Feature Scaling

```{r scaling, echo=TRUE}
features <- features %>%
  mutate(
    age_scaled     = as.numeric(scale(age)),
    premium_scaled = as.numeric(scale(annual_premium))
  )
```

```{r scaling-table}
features %>%
  summarise(
    `Age (mean)`        = round(mean(age), 1),
    `Age scaled (mean)` = round(mean(age_scaled), 3),
    `Age scaled (sd)`   = round(sd(age_scaled), 3)
  ) %>%
  kable(caption = "Table 3: Effect of Scaling (mean ≈ 0, sd ≈ 1)")
```

---

# Step 4 — Feature Usefulness Check

```{r feature-check}
feature_assoc <- features %>%
  group_by(churned) %>%
  summarise(
    `Avg Age`              = round(mean(age), 1),
    `Avg Premium`          = round(mean(annual_premium), 0),
    `Avg Claim/Premium`    = round(mean(claim_to_premium), 3),
    `% High Value`         = round(mean(high_value) * 100, 1),
    .groups = "drop"
  ) %>%
  mutate(Group = if_else(churned == 1, "Churned", "Retained")) %>%
  select(Group, everything(), -churned)
```

```{r feature-check-table}
feature_assoc %>%
  kable(caption = "Table 4: Feature Averages by Churn Status")
```

```{r feature-check-plot, fig.width=8, fig.height=4,fig.cap="Figure 1: Claim-to-Premium Ratio by Churn Status"}
features %>%
  mutate(Status = if_else(churned == 1, "Churned", "Retained")) %>%
  filter(claim_to_premium < quantile(claim_to_premium, 0.95)) %>%
  ggplot(aes(x = Status, y = claim_to_premium, fill = Status)) +
  geom_boxplot(alpha = 0.75) +
  scale_fill_manual(values = c("Churned"  = "#d7191c",
                                "Retained" = "#2c7bb6")) +
  labs(title = "Claim-to-Premium Ratio by Churn Status",
       x = NULL, y = "Claim / Premium Ratio") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

# Step 5 — Assemble Model-Ready Dataset

```{r assemble, echo=TRUE}
model_data <- features %>%
  select(
    churned,                      # target
    age_band, policy_type,        # categorical
    region, gender,
    claim_to_premium,             # engineered numeric
    premium_per_month, high_value,
    age_scaled, premium_scaled    # scaled numeric
  )
```

```{r assemble-summary}
tibble(
  `Feature Type` = c("Target", "Categorical", "Engineered Numeric",
                     "Scaled Numeric", "Total Columns"),
  Count = c(
    1,
    sum(map_lgl(model_data, is.factor)),
    3,
    2,
    ncol(model_data)
  )
) %>%
  kable(caption = "Table 5: Model-Ready Dataset Composition")
```

---

# Key Findings

- Three categories of features were engineered: **derived numeric**
  (ratios and per-unit metrics), **encoded categorical** (factors and
  one-hot dummies), and **scaled numeric** (standardised to mean 0, sd 1).
- The **feature usefulness check** (Table 4) shows churned and retained
  customers differ across the engineered features — confirming they carry
  predictive signal before any model is fitted.
- **Scaling** matters for distance- and gradient-based models (covered in
  Lesson 23); tree-based models are scale-invariant and do not require it.
- The assembled **model-ready dataset** (`r ncol(model_data)` columns)
  is the input for **Lesson 22**, where we fit a logistic regression /
  GLM to predict churn.

---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
