---
title: "Retention Concepts & Churn Data Structures"
subtitle: "Churn & Retention Modelling — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(lubridate)
library(knitr)

policyholders <- read_csv("L2_policyholders.csv") %>%
  mutate(
    policy_start_date = as.Date(policy_start_date),
    cohort_year       = year(policy_start_date),
    churned           = if_else(policy_duration_months < 36, 1, 0)
  )
```

# Overview

This report introduces the core data structures and metrics used in churn
and retention analysis, using the **`r nrow(policyholders)`-policyholder**
portfolio.

Two perspectives are demonstrated: the **snapshot view** (one row per
customer, current status) and the **cohort view** (customers grouped by
acquisition period, tracked over tenure).

Overall churn rate (duration < 36 months):
**`r paste0(round(mean(policyholders$churned) * 100, 1), "%")`**.

---

# Snapshot Data Structure

## One Row Per Customer

```{r snapshot, echo=TRUE}
snapshot <- policyholders %>%
  transmute(
    policyholder_id,
    policy_type,
    cohort_year,
    tenure_months = policy_duration_months,
    status        = if_else(churned == 1, "Churned", "Active")
  )
```

```{r snapshot-table}
snapshot %>%
  head(8) %>%
  kable(caption = "Table 1: Snapshot Data — One Row Per Customer")
```

## Churn Rate by Segment

```{r churn-segment}
snapshot %>%
  group_by(policy_type) %>%
  summarise(
    Customers   = n(),
    Churned     = sum(status == "Churned"),
    `Churn Rate` = paste0(round(mean(status == "Churned") * 100, 1), "%"),
    `Retention Rate` = paste0(round(mean(status == "Active") * 100, 1), "%")
  ) %>%
  kable(caption = "Table 2: Churn & Retention Rate by Policy Type")
```

---

# Cohort Data Structure

## Cohort Sizes by Acquisition Year

```{r cohort-sizes}
policyholders %>%
  count(cohort_year, name = "cohort_size") %>%
  kable(caption = "Table 3: Cohort Size by Acquisition Year")
```

## Retention Matrix

```{r retention-matrix, echo=TRUE}
# For each cohort, compute % still active at tenure milestones
milestones <- c(12, 24, 36, 48)

retention_matrix <- policyholders %>%
  group_by(cohort_year) %>%
  summarise(
    cohort_size = n(),
    `12m` = mean(policy_duration_months >= 12),
    `24m` = mean(policy_duration_months >= 24),
    `36m` = mean(policy_duration_months >= 36),
    `48m` = mean(policy_duration_months >= 48),
    .groups = "drop"
  )
```

```{r retention-matrix-table}
retention_matrix %>%
  mutate(across(ends_with("m"),
                ~ paste0(round(. * 100, 1), "%"))) %>%
  rename(`Cohort Year` = cohort_year,
         `Cohort Size`  = cohort_size) %>%
  kable(caption = "Table 4: Retention Matrix — % Active by Tenure Milestone")
```

---

# Retention Curve Visualisation

```{r retention-curve, fig.width=8, fig.height=5, fig.cap="Figure 1: Retention Curves by Acquisition Cohort"}
retention_long <- retention_matrix %>%
  select(cohort_year, `12m`, `24m`, `36m`, `48m`) %>%
  pivot_longer(-cohort_year,
               names_to  = "tenure",
               values_to = "retention") %>%
  mutate(tenure_months = as.numeric(str_remove(tenure, "m")))

# Add starting point (0 months = 100%)
start_points <- retention_matrix %>%
  transmute(cohort_year, tenure_months = 0, retention = 1)

bind_rows(start_points,
          retention_long %>% select(cohort_year, tenure_months, retention)) %>%
  ggplot(aes(x = tenure_months, y = retention,
             color = factor(cohort_year))) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set2") +
  scale_y_continuous(limits = c(0, 1),
                     labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(0, 48, 12)) +
  labs(title  = "Retention Curves by Acquisition Cohort",
       x = "Tenure (Months)", y = "Retention Rate",
       color = "Cohort Year") +
  theme_minimal(base_size = 12)
```

---

# Churn Rate by Tenure Band

```{r tenure-band, fig.width=8, fig.height=4, fig.cap="Figure 2: Churn Distribution by Tenure Band"}
policyholders %>%
  mutate(tenure_band = cut(policy_duration_months,
                           breaks = c(0, 12, 24, 36, 48, 72),
                           labels = c("0-12", "12-24", "24-36",
                                      "36-48", "48+"))) %>%
  count(tenure_band) %>%
  ggplot(aes(x = tenure_band, y = n, fill = tenure_band)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Customer Count by Tenure Band",
       x = "Tenure Band (Months)", y = "Customers") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

# Key Findings

- The **snapshot view** answers "what is the current state?" — useful for
  reporting headline churn rates by segment.
- The **cohort view** answers "how does retention evolve over tenure?" —
  essential for understanding *when* customers leave, not just *how many*.
- The **retention matrix** (Table 4) is the standard structure for cohort
  analysis: rows are acquisition cohorts, columns are tenure milestones.
  It connects directly to the survival curves from Part V.
- Churn and survival are two views of the same phenomenon: churn rate at
  tenure t equals 1 minus the survival probability S(t).
- In **Lesson 18**, we fit **parametric churn models** (Weibull and
  log-logistic) to these data structures, building on the survival
  modelling foundation from Lessons 15–16.

---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
