---
title: "Forecasting & Diagnostics: Premium Income"
subtitle: "Time Series Forecasting — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(forecast)
library(knitr)

monthly_premiums <- read_csv("L5_monthly_premiums.csv") %>%
  mutate(month = as.Date(month))

premium_ts <- ts(
  monthly_premiums$total_premium,
  start     = c(2019, 1),
  frequency = 12
)

# Train / test split: hold out last 12 months
train_ts <- window(premium_ts, end   = c(2022, 12))
test_ts  <- window(premium_ts, start = c(2023,  1))
```

# Overview

This report produces a 12-month forward forecast from the ARIMA model
fitted in Lesson 8, and evaluates forecast accuracy using a
**train/test split** — holding out the final
**`r length(test_ts)` months** of the series as unseen test data.

The series spans
**`r format(min(monthly_premiums$month), '%B %Y')`** to
**`r format(max(monthly_premiums$month), '%B %Y')`**
(**`r length(premium_ts)`** observations total).

---

# Train / Test Split

```{r split-plot, echo=TRUE, fig.width=8, fig.height=4,fig.cap="Figure 1: Train/Test Split of Premium Income Series"}
autoplot(train_ts, series = "Train") +
  autolayer(test_ts, series = "Test") +
  scale_color_manual(values = c("Train" = "#2c7bb6", "Test" = "#d7191c")) +
  labs(title = "Premium Income — Train / Test Split",
       x = NULL, y = "Total Premium (£)", color = NULL) +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  theme_minimal(base_size = 12)
```

---

# Fit Model on Training Data

```{r fit-train, echo=TRUE}
model_train <- auto.arima(
  train_ts,
  stepwise      = FALSE,
  approximation = FALSE
)

ord <- arimaorder(model_train)
```

Model selected on training data:
**ARIMA(`r ord["p"]`,`r ord["d"]`,`r ord["q"]`)(`r ord["P"]`,`r ord["D"]`,`r ord["Q"]`)[12]**

---

# Forecast

## 12-Month Ahead Forecast

```{r forecast-obj, echo=TRUE}
fc <- forecast(model_train, h = 12, level = c(80, 95))
```

```{r forecast-plot, echo=FALSE, fig.width=9, fig.height=5,fig.cap="Figure 2: 12-Month Ahead Forecast with Prediction Intervals"}
autoplot(fc) +
  autolayer(test_ts, series = "Actual", color = "#d7191c") +
  labs(title = "12-Month Forecast vs Actual",
       x = NULL, y = "Total Premium (£)") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  theme_minimal(base_size = 12)
```

## Forecast Values Table

```{r forecast-table}
fc_df <- as.data.frame(fc) %>%
  mutate(Month = seq(as.Date("2023-01-01"),
                     by = "month", length.out = 12),
         Actual = as.numeric(test_ts)) %>%
  select(Month,
         `Point Forecast`,
         `Lo 80`, `Hi 80`,
         `Lo 95`, `Hi 95`,
         Actual) %>%
  mutate(across(where(is.numeric), ~ round(., 0)))

fc_df %>%
  kable(caption = "Table 1: Forecast vs Actual — Monthly Detail")
```

---

# Forecast Accuracy

## Out-of-Sample Accuracy

```{r accuracy-table}
acc <- accuracy(fc, test_ts)

as.data.frame(acc) %>%
  rownames_to_column("Set") %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kable(caption = "Table 2: Forecast Accuracy — Training and Test Sets")
```

## Comparison with Naïve Benchmark

```{r naive-comparison, echo=TRUE}
model_naive  <- snaive(train_ts, h = 12)
acc_arima    <- accuracy(fc,           test_ts)
acc_naive    <- accuracy(model_naive,  test_ts)
```

```{r benchmark-table}
tibble(
  Model = c("ARIMA (auto)", "Seasonal Naïve"),
  RMSE  = round(c(acc_arima["Test set",  "RMSE"],
                   acc_naive["Test set",  "RMSE"]), 0),
  MAE   = round(c(acc_arima["Test set",  "MAE"],
                   acc_naive["Test set",  "MAE"]),  0),
  MAPE  = round(c(acc_arima["Test set",  "MAPE"],
                   acc_naive["Test set",  "MAPE"]),  2)
) %>%
  kable(caption = "Table 3: ARIMA vs Seasonal Naïve — Test Set Accuracy")
```

---

# Full-Series Forecast (Forward-Looking)

```{r full-model}
model_full <- auto.arima(
  premium_ts,
  stepwise      = FALSE,
  approximation = FALSE
)

fc_full <- forecast(model_full, h = 12, level = c(80, 95))
```

```{r full-forecast-plot, echo=FALSE, fig.width=9, fig.height=5,fig.cap="Figure 3: Forward-Looking 12-Month Forecast from Full Series"}
autoplot(fc_full) +
  labs(title = "12-Month Forward Forecast — Full Series",
       x = NULL, y = "Total Premium (£)") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  theme_minimal(base_size = 12)
```

---

# Key Findings

```{r kf-inline, include=FALSE}
rmse_arima <- round(acc_arima["Test set", "RMSE"], 0)
rmse_naive <- round(acc_naive["Test set", "RMSE"], 0)
mape_arima <- round(acc_arima["Test set", "MAPE"], 2)
```

- The ARIMA model achieved a test-set RMSE of
  **£`r format(rmse_arima, big.mark=",")`** and MAPE of
  **`r mape_arima`%** on the held-out 12 months.
- The seasonal naïve benchmark had RMSE of
  **£`r format(rmse_naive, big.mark=",")`** —
  `r ifelse(rmse_arima < rmse_naive, "the ARIMA model outperforms the naïve benchmark.", "the naïve benchmark is competitive — further model refinement may be warranted.")`
- The 80% and 95% prediction intervals widen over the forecast horizon,
  correctly reflecting increasing uncertainty over time.
- The forward-looking forecast (Figure 3) provides the 12-month premium
  income projection that would feed directly into a budgeting or
  reserving exercise.

---

*Report generated automatically using R Markdown.
Data source: `monthly_premiums.csv`.*
