---
title: "Time Series & Retention Curve Visualisation"
subtitle: "Exploratory Visualisation — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(forecast)
library(knitr)

policyholders  <- read_csv("policyholders.csv")
monthly_premiums <- read_csv("monthly_premiums.csv") %>%
  mutate(month = as.Date(month))
```

# Overview

This report visualises two key actuarial data structures: a **monthly premium
income time series** spanning `r format(min(monthly_premiums$month), '%B %Y')`
to `r format(max(monthly_premiums$month), '%B %Y')`, and **policyholder
retention curves** derived from the portfolio dataset.

---

# Premium Income Time Series

## Monthly Premium Trend

```{r ts-object, echo=TRUE}
premium_ts <- ts(
  monthly_premiums$total_premium,
  start     = c(2019, 1),
  frequency = 12
)
```

```{r autoplot, echo=TRUE, fig.width=8, fig.height=4,
    fig.cap="Figure 1: Monthly Premium Income (Jan 2019 – Dec 2023)"}
autoplot(premium_ts) +
  labs(title = "Monthly Premium Income",
       x     = "Date",
       y     = "Total Premium (£)") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  theme_minimal(base_size = 12)
```

## ggplot Version with Date Axis

```{r ggplot-ts, echo=TRUE, fig.width=8, fig.height=4,
    fig.cap="Figure 2: Monthly Premium Income with Formatted Date Axis"}
ggplot(monthly_premiums, aes(x = month, y = total_premium)) +
  geom_line(color = "#2c7bb6", linewidth = 0.8) +
  geom_smooth(method = "lm", se = FALSE,
              color = "#d7191c", linetype = "dashed", linewidth = 0.6) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  labs(title   = "Monthly Premium Income with Trend Line",
       x = NULL, y = "Total Premium (£)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

# Retention Curves

## Empirical Retention by Policy Type

```{r retention-data, echo=TRUE}
retention <- policyholders %>%
  group_by(policy_type, policy_duration_months) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(policy_type, policy_duration_months) %>%
  group_by(policy_type) %>%
  mutate(
    total        = sum(n),
    survivors    = total - cumsum(lag(n, default = 0)),
    retention_pct = survivors / total * 100
  )
```

```{r retention-plot, echo=TRUE, fig.width=8, fig.height=5,
    fig.cap="Figure 3: Empirical Retention Curves by Policy Type"}
ggplot(retention, aes(x = policy_duration_months,
                      y = retention_pct,
                      color = policy_type)) +
  geom_step(linewidth = 0.8) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(0, 100),
                     labels = scales::percent_format(scale = 1)) +
  scale_x_continuous(breaks = seq(0, 72, 12)) +
  labs(title  = "Empirical Retention Curves by Policy Type",
       x      = "Policy Duration (Months)",
       y      = "Retention (%)",
       color  = "Policy Type") +
  theme_minimal(base_size = 12)
```

## Retention Summary at Key Durations

```{r retention-table}
retention %>%
  filter(policy_duration_months %in% c(12, 24, 36, 48)) %>%
  select(policy_type, policy_duration_months, retention_pct) %>%
  mutate(retention_pct = paste0(round(retention_pct, 1), "%")) %>%
  pivot_wider(names_from  = policy_duration_months,
              values_from = retention_pct,
              names_prefix = "Month ") %>%
  kable(caption = "Table 1: Retention Rate at Key Durations by Policy Type")
```

---

# Key Findings

- Premium income shows a clear **upward trend** over the observation period,
  with seasonal variation in winter months.
- Retention curves differ by product line — some products show steeper
  early-duration drop-off, which has implications for pricing and reserving.
- The 12-month and 24-month retention rates are particularly relevant for
  IFoA CS2 lapse modelling assumptions.

---

*Report generated automatically using R Markdown.
Data sources: `policyholders.csv`, `monthly_premiums.csv`.*
