---
title: "Logistic Regression & GLM Churn Prediction"
subtitle: "Applied Predictive & Causal Analytics — Actuarial Analytics Division"
author: "Actuarial Analytics Team"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(tidyverse)
library(broom)
library(pROC)
library(knitr)

set.seed(42)

policyholders <- read_csv("L2_policyholders.csv") %>%
  mutate(
    churned     = if_else(policy_duration_months < 24, 1, 0),
    age_band    = cut(age, breaks = c(0, 35, 50, 65, 100),
                      labels = c("Under35", "Age35_50", "Age50_65", "Over65")),
    policy_type = factor(policy_type),
    region      = factor(region),
    claim_to_premium = claim_amount / annual_premium
  )

# Train / test split (70/30)
train_idx <- sample(nrow(policyholders), 0.7 * nrow(policyholders))
train <- policyholders[train_idx, ]
test  <- policyholders[-train_idx, ]
```

# Overview

This report fits a **logistic regression GLM** to predict policyholder
churn, interprets the fitted coefficients as odds ratios, and evaluates
classification performance on a held-out test set.

Data: **`r nrow(policyholders)`** policyholders
(**`r nrow(train)`** train / **`r nrow(test)`** test).
Base churn rate: **`r paste0(round(mean(policyholders$churned)*100,1),"%")`**.

---

# Fit the Logistic Regression Model

```{r fit-glm, echo=TRUE}
model <- glm(
  churned ~ age_band + policy_type + annual_premium +
            claim_to_premium + region,
  data   = train,
  family = binomial(link = "logit")
)
```

```{r model-summary, echo=TRUE}
tidy(model) %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  kable(caption = "Table 1: GLM Coefficient Estimates (log-odds scale)")
```

---

# Interpreting Coefficients as Odds Ratios

```{r odds-ratios, echo=TRUE}
odds_ratios <- tidy(model) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    odds_ratio = exp(estimate),
    effect     = if_else(odds_ratio > 1,
                         "Increases churn odds",
                         "Decreases churn odds")
  ) %>%
  select(term, estimate, odds_ratio, p.value, effect) %>%
  arrange(desc(odds_ratio))
```

```{r odds-ratio-table}
odds_ratios %>%
  mutate(estimate   = round(estimate, 3),
         odds_ratio = round(odds_ratio, 3),
         p.value    = round(p.value, 4)) %>%
  rename(Term = term, `Log-Odds` = estimate,
         `Odds Ratio` = odds_ratio, `p-value` = p.value,
         Effect = effect) %>%
  kable(caption = "Table 2: Odds Ratios (sorted, highest churn effect first)")
```

An odds ratio above 1 means that predictor increases the odds of churn
relative to the baseline category; below 1 means it decreases them.

---

# Predictions on the Test Set

```{r predict, echo=TRUE}
test <- test %>%
  mutate(pred_prob = predict(model, newdata = test, type = "response"))
```

With an imbalanced target (churn is the minority class), the default
0.5 threshold often classifies *every* case as the majority class.
We therefore select the threshold that maximises **Youden's J statistic**
(sensitivity + specificity − 1) from the ROC curve.

```{r threshold, echo=TRUE}
roc_obj   <- roc(test$churned, test$pred_prob, quiet = TRUE)
auc_val   <- auc(roc_obj)

# Optimal threshold by Youden's J
opt       <- coords(roc_obj, "best", best.method = "youden",
                    ret = "threshold", transpose = FALSE)
threshold <- as.numeric(opt$threshold[1])

test <- test %>%
  mutate(pred_class = if_else(pred_prob > threshold, 1, 0))

cat("Base churn rate:    ", round(mean(test$churned), 3), "\n")
cat("Chosen threshold:   ", round(threshold, 3), "\n")
cat("Predicted churn rate:", round(mean(test$pred_class), 3), "\n")
```

```{r pred-dist-plot, fig.width=8, fig.height=4,fig.cap="Figure 1: Predicted Churn Probability by Actual Status"}
test %>%
  mutate(Actual = if_else(churned == 1, "Churned", "Retained")) %>%
  ggplot(aes(x = pred_prob, fill = Actual)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = threshold, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values = c("Churned"  = "#d7191c",
                                "Retained" = "#2c7bb6")) +
  labs(title = "Predicted Churn Probability by Actual Status",
       subtitle = paste0("Dashed line = chosen threshold (",
                         round(threshold, 3), ")"),
       x = "Predicted Probability", y = "Count") +
  theme_minimal(base_size = 12)
```

---

# Classification Performance

## Confusion Matrix

```{r confusion}
# Force both classes (0/1) to appear even if one is empty
conf <- table(
  Predicted = factor(test$pred_class, levels = c(0, 1)),
  Actual    = factor(test$churned,    levels = c(0, 1))
)

as.data.frame.matrix(conf) %>%
  rownames_to_column("Predicted") %>%
  rename(`Actual: Retained (0)` = `0`,
         `Actual: Churned (1)`  = `1`) %>%
  kable(caption = paste0("Table 3: Confusion Matrix (threshold = ",
                         round(threshold, 3), ")"))
```

## Performance Metrics

```{r metrics}
TP <- sum(test$pred_class == 1 & test$churned == 1)
TN <- sum(test$pred_class == 0 & test$churned == 0)
FP <- sum(test$pred_class == 1 & test$churned == 0)
FN <- sum(test$pred_class == 0 & test$churned == 1)

safe_div  <- function(a, b) if (b == 0) NA_real_ else a / b

accuracy  <- (TP + TN) / nrow(test)
precision <- safe_div(TP, TP + FP)
recall    <- safe_div(TP, TP + FN)
f1        <- safe_div(2 * precision * recall, precision + recall)

tibble(
  Metric = c("Accuracy", "Precision", "Recall (Sensitivity)", "F1 Score"),
  Value  = round(c(accuracy, precision, recall, f1), 3)
) %>%
  kable(caption = "Table 4: Classification Performance Metrics")
```

## ROC Curve & AUC

```{r roc, echo=TRUE, fig.width=6.5, fig.height=5,fig.cap="Figure 2: ROC Curve"}
# roc_obj and auc_val already computed in the threshold chunk above
tibble(
  fpr = 1 - roc_obj$specificities,
  tpr = roc_obj$sensitivities
) %>%
  ggplot(aes(x = fpr, y = tpr)) +
  geom_line(color = "#2c7bb6", linewidth = 0.9) +
  geom_abline(linetype = "dashed", color = "grey50") +
  annotate("text", x = 0.6, y = 0.2,
           label = paste0("AUC = ", round(auc_val, 3)),
           size = 5, color = "#d7191c") +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "ROC Curve — Churn GLM",
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal(base_size = 12)
```

---

# Key Findings

```{r kf-vals, include=FALSE}
auc_round <- round(as.numeric(auc_val), 3)
acc_round <- round(accuracy, 3)
```

- The GLM achieves a test-set **accuracy of `r acc_round`** and
  **AUC of `r auc_round`** — AUC above 0.5 confirms the model
  discriminates churners from non-churners better than chance.
- **Threshold choice matters under class imbalance.** A naive 0.5
  cut-off can classify every case as the majority class, producing a
  one-row confusion matrix. Selecting the threshold by Youden's J
  restores a meaningful churn/retain split.
- **Odds ratios** (Table 2) provide directly interpretable effects:
  each predictor's multiplicative impact on the odds of churn. This
  interpretability is why GLMs remain a regulatory and actuarial standard.
- **Recall** is the most business-relevant metric for churn — it measures
  the proportion of actual churners the model correctly flags, which
  drives retention-campaign coverage.
- In **Lesson 23**, we compare this GLM against tree-based machine
  learning models (Random Forest, XGBoost) to assess whether added
  complexity improves predictive performance.

---

*Report generated automatically using R Markdown.
Data source: `policyholders.csv`.*
